---
title: "eClip_analysis_repeat"
author: "Giorgia Bucciarelli"
date: "2023-09-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, root.dir = "~/Desktop/YBX1_giorgia/eClip/")
#setwd("~/Desktop/YBX1")
```

```{r, warning=F, message=F, echo=FALSE}
library(tidyverse)
library(data.table)
library(ggrepel) #extra geoms for ggplot2
library(ggbeeswarm) #to create categorical scatter plots (violin plots)
library(ggthemes) #extra themes, scales and geoms for ggplot2
library(ggseqlogo)
library(factoextra)

library(cliProfiler)
library(GenomicRanges) # to deal with genomic intervals (BED files)
library(BSgenome) # genome functions
library(BSgenome.Hsapiens.UCSC.hg38) # human genome
library(Biostrings) # deals with sequences (e.g. fasta files), alphabetfrequency
library(umap)
library(edgeR) #to perform normalization
library(ggbreak)
library(hrbrthemes)
library(enrichR)
#library(memes) # to perform motif discovery

library(wesanderson)
library(extrafont)
loadfonts(quiet = T)
bs <- 11
bf <- "Arial"
```

```{r}
ACGU_color <- c("#0f9648","#255c99","#f7b32b","#d62839")
```

# eClip analysis

```{r resume_bedcoverage, warning=F,message=F,echo=F, eval=T}

#function to read the bed files

resume_bedcoverage<-function(dir, # directory with the bed files
                             name_cols=c(1,2,3,6), # numeric or vector: columns with feature names, coordinate dei picchi
                             cov_col=7   # numeric: column with coverage values
) {
   
   list_input<-fs::dir_ls(dir,glob="*.bed",type="file") #
   
   if(exists("o_tab")){
      rm(o_tab)
   }
   
   for(i in (list_input)){
      
      i_tab<-read_tsv(i,col_names=F, progress = F,show_col_types = F)
      
      s_name<-str_split(i,dir,simplify=T)[,2] 
      s_name<-str_split(s_name,"\\.",simplify=T)[,1]
      
      x_tab <- i_tab %>% unite(id_name,all_of(name_cols)) %>% dplyr::select(id_name)
      x_tab[,s_name] <- i_tab[,cov_col]
      
      rm(i_tab)
      
      if(!exists("o_tab")){
         o_tab<- x_tab
      } else {
         o_tab<- full_join(o_tab,x_tab,by="id_name")
      }
   }
   return(o_tab)
}

```

## Bed files loading 

```{r load BED files}
same_cov <- resume_bedcoverage(dir = "./peaks_cov_pureclip/same/", 
                               name_cols = c(1,2,3,6),
                               cov_col = 7)

```

```{r merge peaks, warning=F, message=F, echo=F, eval=FALSE}
# dataframe of number of peaks in each chr region for each sample
## chr regions on rows, samples on columns

counts_raw_df <- as.data.frame(same_cov)
rownames(counts_raw_df) <- counts_raw_df$id_name
counts_raw_df$id_name <- NULL

colnames(counts_raw_df) <- str_sub(colnames(counts_raw_df), 1, -15) 

```

```{r c_anno, warning=F, message=F, echo=F, eval=FALSE}

c_anno <- data.frame("sample" = colnames(counts_raw_df))
rownames(c_anno) <- c_anno$sample

c_anno_add <- as.data.frame(str_split(string = c_anno$sample, pattern = "_", simplify = T))
colnames(c_anno_add) <- c("library", "cell", "replicate")

c_anno <- cbind(c_anno, c_anno_add) #column bind the two dfs
c_anno$condition <- paste(c_anno$library, c_anno$cell, sep = "_")
c_anno$condition <- factor(c_anno$condition)

counts_raw_df <- counts_raw_df[,c_anno$sample]

rm(c_anno_add)
```


```{r filter_dt}

counts_raw_dt <- as.data.table(counts_raw_df)
counts_raw_dt$id <- rownames(counts_raw_df)

c_anno_dt <- as.data.table(c_anno)

sig_thr <- 10 #count threshold

counts_long_dt <- counts_raw_dt[,id:=as.factor(id)]
counts_long_dt <- melt.data.table(counts_long_dt, id.vars = c("id"), variable.name = "sample", value.name = "counts")#wide to long reshaping tool, similar to pivot_longer?

setkey(counts_long_dt, sample) #sorts a data.table and marks it as sorted, the sorted columns are the key, always in ascending order
setkey(c_anno_dt, sample)

counts_long_dt <- counts_long_dt[c_anno_dt, on = "sample"]

counts_filter_dt <- counts_long_dt[,.(min_c = min(counts)), .(id, condition)][condition %in% c("IP_SCC", "IP_HK")]
counts_filter_dt <- counts_filter_dt[,.(passes_c = sum(min_c >= sig_thr)), .(id)] 
filter_id <- as.character(counts_filter_dt[passes_c > 0]$id)

counts_df <- counts_raw_df[filter_id,]
# filter to retain regions with a count greater than 10

```

```{r}
rm(counts_long_dt,
   counts_filter_dt, 
   counts_raw_dt, 
   c_anno_dt)
```

## Peak Annotation

```{r}
#load("Results/filtered_modification_df.RData")
load("~/Desktop/anno/ens_anno_110_hs.RData")

# creating exon and gene annotation
ense_slim <- ense |> dplyr::select(ensembl_exon_id,
                                   rank, 
                                   exon_chrom_start,
                                   '5_utr_start',
                                   "5_utr_end",
                                   genomic_coding_start,
                                   genomic_coding_end,
                                   "3_utr_start",
                                   "3_utr_end",
                                   exon_chrom_end,
                                   l_exon,
                                   l_5utr,
                                   l_3utr,
                                   l_cds,
                                   TIS, TTS, exon_class,
                                   ensembl_transcript_id) 


# transcript annotation
enst_slim <- enst |> dplyr::select(external_transcript_name,
                                   transcript_biotype, transcript_length,
                                   canonical, ensembl_transcript_id, ensembl_gene_id)

## We filter to retain only protein coding transcripts for protein coding genes
# to avoid annotation override based on other biotypes
enst_filt <- enst_slim |> 
  group_by(ensembl_gene_id, transcript_biotype) |> 
  summarise(N = n(), transcripts = paste(external_transcript_name, collapse = " "))
enst_filt$coding <- ifelse(enst_filt$transcript_biotype == "protein_coding", T, F)

coding_ense <- enst_filt |> filter(coding == T) |> pull(ensembl_gene_id) |> unique()
enst_filt_n_cod <- enst_filt |> dplyr::filter(ensembl_gene_id %in% coding_ense & transcript_biotype != "protein_coding")
excluded_t <- unlist(strsplit(enst_filt_n_cod$transcripts, " "))

enst_slim_cod <- enst_slim |> dplyr::filter(!external_transcript_name %in% excluded_t)
ense_slim_cod <- ense_slim |> dplyr::filter(ensembl_transcript_id %in% enst_slim_cod$ensembl_transcript_id) 

rownames(enst_slim) <- enst_slim$ensembl_transcript_id
enst_slim <- enst_slim[ense_slim$ensembl_transcript_id,]
enst_slim$ensembl_transcript_id <- NULL

# gene annotation
ensg_slim <- ensg |> dplyr::select(external_gene_name, ensembl_gene_id, gene_biotype, chr_ucsc, strand_ucsc)
rownames(ensg_slim) <- ensg_slim$ensembl_gene_id
ensg_slim <- ensg_slim[enst_slim$ensembl_gene_id,]
ensg_slim$ensembl_gene_id <- NULL

# single annotation 
ense_df <- left_join(enst_slim_cod, ensg, by = "ensembl_gene_id")
ense_df <- left_join(ense_df, ense_slim_cod, by = "ensembl_transcript_id")

#ense_df <- cbind(ense_slim_cod, ensg_slim) #combines all the annotation dataframes into one
# in rownames we have the gene names
#rownames(ense_df)<-make.unique(str_c(ense_df$external_transcript_name, ense_df$rank, sep="_")) #combines external transcript name and rank with "_"


# the following function takes a dataframe-like object as input and tries to automatically find the columns that describe genomic ranges
# returns a GRanges object
ense_granges <- makeGRangesFromDataFrame(ense_df,
                                         keep.extra.columns = T, 
                                         ignore.strand = F,
                                         seqinfo = NULL,
                                         seqnames.field = "chr_ucsc", 
                                         start.field = "exon_chrom_start",  
                                         end.field = "exon_chrom_end", 
                                         strand.field = "strand_ucsc",
                                         starts.in.df.are.0based = F)

ensg_df <- ensg |> dplyr::select(external_gene_name, ensembl_gene_id, gene_biotype, chr_ucsc, strand_ucsc, start_position, end_position, description)

ensg_granges <- makeGRangesFromDataFrame(ensg_df,
                                         keep.extra.columns = T, 
                                         ignore.strand = F,
                                         seqinfo = NULL,
                                         seqnames.field = "chr_ucsc", 
                                         start.field = "start_position",  
                                         end.field = "end_position", 
                                         strand.field = "strand_ucsc",
                                         starts.in.df.are.0based = F)

```

```{r}
## 5' UTR granges
ense_5_utr_granges <- makeGRangesFromDataFrame(ense_df |> dplyr::filter(!is.na(l_5utr)),
                                               keep.extra.columns = T,
                                               ignore.strand = F,
                                               seqinfo = NULL,
                                               seqnames.field = "chr_ucsc",
                                               start.field = "5_utr_start",
                                               end.field = "5_utr_end",
                                               strand.field = "strand_ucsc",
                                               starts.in.df.are.0based = F)



## CDS granges
ense_cds_granges <- makeGRangesFromDataFrame(ense_df |> dplyr::filter(!is.na(l_cds)),
                                               keep.extra.columns = T,
                                               ignore.strand = F,
                                               seqinfo = NULL,
                                               seqnames.field = "chr_ucsc",
                                               start.field = "genomic_coding_start",
                                               end.field = "genomic_coding_end",
                                               strand.field = "strand_ucsc",
                                               starts.in.df.are.0based = F)

## 3' UTR granges
ense_3_utr_granges <- makeGRangesFromDataFrame(ense_df |> dplyr::filter(!is.na(l_3utr)),
                                               keep.extra.columns = T,
                                               ignore.strand = F,
                                               seqinfo = NULL,
                                               seqnames.field = "chr_ucsc",
                                               start.field = "3_utr_start",
                                               end.field = "3_utr_end",
                                               strand.field = "strand_ucsc",
                                               starts.in.df.are.0based = F)

```


```{r}
r_anno <- data.frame("id_name" = rownames(counts_df))
r_anno_add <- as.data.frame(str_split(string = r_anno$id_name, pattern = "_", simplify = T))
colnames(r_anno_add) <- c("chr", "start", "end", "strand")

r_anno <- cbind(r_anno, r_anno_add)
r_anno$start <- as.numeric(r_anno$start)
r_anno$end <- as.numeric(r_anno$end)
r_anno$length <- r_anno$end - r_anno$start

chr_conversion_canonical <- data.frame(
   "chr_ens"=c("1","2","3","4","5","6","7","8","9","10",
      "11","12","13","14","15","16","17","18","19","20",
      "21","22","X","Y","MT"),
   "chr_ucsc"=c("chr1","chr2","chr3","chr4","chr5","chr6","chr7","chr8","chr9","chr10",
      "chr11","chr12","chr13","chr14","chr15","chr16","chr17","chr18","chr19","chr20",
      "chr21","chr22","chrX","chrY","chrM"))

r_anno <- r_anno |> dplyr::filter(chr %in% chr_conversion_canonical$chr_ucsc) #keep only the canonical chromosomes

# check for peak width
peak_granges <- makeGRangesFromDataFrame(r_anno,
                                         keep.extra.columns = F,
                                         ignore.strand = F,
                                         seqinfo = NULL,
                                         seqnames.field = "chr",
                                         start.field = "start",
                                         end.field = "end",
                                         strand.field = "strand",
                                         starts.in.df.are.0based = T)

r_anno$peak_seq <- getSeq(BSgenome.Hsapiens.UCSC.hg38, peak_granges) |> as.character() |> stringr::str_replace_all("T", "U")

r_anno <- r_anno[!(str_detect(r_anno$peak_seq, "N")),] #remove sequences with N

peak_granges <- makeGRangesFromDataFrame(r_anno,
                                         keep.extra.columns = T,
                                         ignore.strand = F,
                                         seqinfo = NULL,
                                         seqnames.field = "chr",
                                         start.field = "start",
                                         end.field = "end",
                                         strand.field = "strand",
                                         starts.in.df.are.0based = T)

counts_df <- counts_df[r_anno$id_name,]
```

```{r}
## Annotation on genes to find intergenic regions and distinguish them from those spanning genes

over_genes<- GenomicRanges::findOverlaps(peak_granges, ensg_granges) #this function returns the overlap (n.rows) between peaks and genes

### Additional ----
## indentifica le regioni di sovrapposizione
genes_overlaps_ranges <- pintersect(peak_granges[queryHits(over_genes)], ensg_granges[subjectHits(over_genes)])

## calcola la larghezza delle sovrapposizioni
#genes_overlap_widths <- width(genes_overlaps_ranges)


peak_2_ensg <- peak_granges[queryHits(over_genes)] %>% as_tibble() %>% 
  dplyr::select(-width)
ensg_2_peak<- ensg_granges[subjectHits(over_genes)] %>% as_tibble() %>%
  dplyr::rename(gene_chrom_start=start,
                gene_chrom_end=end,
                gene_width=width) %>% 
  dplyr::select(-seqnames,-strand)

peak_ensg_df <-bind_cols(peak_2_ensg,ensg_2_peak) 
#mod_ensg_df$overlap_width <- genes_overlap_widths
#peak_ensg_df <- peak_ensg_df %>% dplyr::filter(!(id_name %in% peak_ense_add$id_name))

peak_ensg_df$region_class <- "Intron"
peak_ensg_df$gene_biotype<-factor(peak_ensg_df$gene_biotype)
peak_ensg_df$gene_biotype<-relevel(peak_ensg_df$gene_biotype, ref = "protein_coding")

peak_ensg_dt<- as.data.table(peak_ensg_df)
setkey(peak_ensg_dt,gene_biotype,gene_width,external_gene_name)
peak_ensg_dt<-peak_ensg_dt[order(gene_biotype,gene_width,external_gene_name)]
peak_ensg_unique <- peak_ensg_dt[!duplicated(id_name)]
```

```{r}
r_anno <- left_join(r_anno, peak_ensg_unique |> dplyr::select(id_name, external_gene_name))

r_anno$region_class <- "Intergenic"
r_anno$region_class[which(r_anno$id_name %in% peak_ensg_unique$id_name)] <- "Intron"
```

```{r}
rm(over_genes,genes_overlaps_ranges,peak_2_ensg,ensg_2_peak, peak_ensg_df, peak_ensg_dt)
```


```{r}
## Annotation of peaks over exonic regions (still no details about 5', CDS, 3')
over_exons <- GenomicRanges::findOverlaps(peak_granges, ense_granges)

## indentifica le regioni di sovrapposizione
exons_overlaps_ranges <- pintersect(peak_granges[queryHits(over_exons)], ense_granges[subjectHits(over_exons)])

## calcola la larghezza delle sovrapposizioni
#exons_overlap_widths <- width(exons_overlaps_ranges)

peak_2_ense <- peak_granges[queryHits(over_exons)] %>% as_tibble() %>% 
  dplyr::select(-width)
ense_2_peak<- ense_granges[subjectHits(over_exons)] %>% as_tibble() %>%
  dplyr::rename(exon_chrom_start=start,
                exon_chrom_end=end,
                exon_width=width) %>% 
  dplyr::select(-seqnames,-strand)

peak_ense_df <-bind_cols(peak_2_ense,ense_2_peak) 
#mod_ense_df$overlap_width <- exons_overlap_widths
#peak_ensg_df <- peak_ensg_df %>% dplyr::filter(!(id_name %in% peak_ense_add$id_name))

peak_ense_df$region_class <- "Non-coding Exon"
peak_ense_df$gene_biotype<-factor(peak_ense_df$gene_biotype)
peak_ense_df$gene_biotype<-relevel(peak_ense_df$gene_biotype, ref = "protein_coding")

peak_ense_dt<- as.data.table(peak_ense_df)
setkey(peak_ense_dt,gene_biotype,exon_width,external_gene_name)
peak_ense_dt<-peak_ense_dt[order(gene_biotype,exon_width,external_gene_name)]
peak_ense_unique <- peak_ense_dt[!duplicated(id_name)]

r_anno$region_class[which(r_anno$id_name %in% peak_ense_unique$id_name)] <- "Non-coding Exon"

## Add annotations of the exons 
# peak_anno <- peak_ensg_unique
# peak_anno$region_class[which(peak_ensg_unique$id_name %in% peak_ense_unique$id_name)] <- peak_ense_unique$region_class
# 
peak_anno_ense_eng <- full_join(peak_ense_unique, peak_ensg_unique)
peak_anno_ense_ensg_dt<- as.data.table(peak_anno_ense_eng)
setkey(peak_anno_ense_ensg_dt,gene_biotype,exon_width,external_gene_name)
peak_anno_ense_ensg_dt<- peak_anno_ense_ensg_dt[order(gene_biotype,exon_width,external_gene_name)]
peak_anno_ense_ensg_unique <- peak_anno_ense_ensg_dt[!duplicated(id_name)]
```

```{r}
rm(over_exons,exons_overlaps_ranges,peak_2_ense,ense_2_peak,peak_ense_df,peak_ense_dt,peak_anno_ense_eng,peak_anno_ense_ensg_dt)
```

```{r over 5'utr}
over_5_utr <- GenomicRanges::findOverlaps(peak_granges, ense_5_utr_granges) ## overlap between peaks and 5'UTR

overlap_ranges_5utr <- pintersect(peak_granges[queryHits(over_5_utr)], ense_5_utr_granges[subjectHits(over_5_utr)])
#overlap_widths_5utr <- width(overlap_ranges_5utr)

peak_2_5utr <- peak_granges[queryHits(over_5_utr)] |> as.tibble() #|> dplyr::rename(_width = width)
ense_5utr_2_peak <- ense_5_utr_granges[subjectHits(over_5_utr)] |> as.tibble() |> dplyr::select(-width, -seqnames, -strand) |> dplyr::rename(start_5_utr = start, end_5_utr = end)

peak_5utr_df <- bind_cols(peak_2_5utr, ense_5utr_2_peak)
#mod_base_5utr_df$overlap_width <- overlap_widths_5utr
peak_5utr_df$region_class <- "5_UTR"

peak_5utr_dt<- as.data.table(peak_5utr_df)
setkey(peak_5utr_dt, canonical, region_class, external_transcript_name, rank)
peak_5utr_dt<-peak_5utr_dt[order(canonical,exon_class,external_transcript_name,rank)]
peak_5utr_unique <- peak_5utr_dt[!duplicated(id_name)]

r_anno$region_class[which(r_anno$id %in% peak_5utr_unique$id_name)] <- "5_UTR"
```

```{r}
rm(over_5_utr, overlap_ranges_5utr, peak_2_5utr, ense_5utr_2_peak, peak_5utr_df,peak_5utr_dt)
```

```{r}
over_cds <- GenomicRanges::findOverlaps(peak_granges, ense_cds_granges) ## overlap between peaks and 5'UTR

overlap_ranges_cds <- pintersect(peak_granges[queryHits(over_cds)], ense_cds_granges[subjectHits(over_cds)])
#overlap_widths_5utr <- width(overlap_ranges_5utr)

peak_2_cds <- peak_granges[queryHits(over_cds)] |> as.tibble() #|> dplyr::rename(_width = width)
ense_cds_2_peak <- ense_cds_granges[subjectHits(over_cds)] |> as.tibble() |> dplyr::select(-width, -seqnames, -strand) |> dplyr::rename(start_cds = start, end_cds = end)

peak_cds_df <- bind_cols(peak_2_cds, ense_cds_2_peak)
#peak_5utr_df$overlap_width <- overlap_widths_5utr
peak_cds_df$region_class <- "CDS"

peak_cds_dt<- as.data.table(peak_cds_df)
setkey(peak_cds_dt, canonical, region_class, external_transcript_name, rank)
peak_cds_dt<-peak_cds_dt[order(canonical,exon_class,external_transcript_name,rank)]
peak_cds_unique <- peak_cds_dt[!duplicated(id_name)]

r_anno$region_class[which(r_anno$id_name %in% peak_cds_unique$id_name)] <- "CDS"
```

```{r}
rm(over_cds, overlap_ranges_cds, peak_2_cds,ense_cds_2_peak,peak_cds_df,peak_cds_dt)
```

```{r over 3'utr}
over_3_utr <- GenomicRanges::findOverlaps(peak_granges, ense_3_utr_granges) ## overlap between peaks and 5'UTR

overlap_ranges_3utr <- pintersect(peak_granges[queryHits(over_3_utr)], ense_3_utr_granges[subjectHits(over_3_utr)])
#overlap_widths_5utr <- width(overlap_ranges_5utr)

peak_2_3utr <- peak_granges[queryHits(over_3_utr)] |> as.tibble() #|> dplyr::rename(_width = width)
ense_3utr_2_peak <- ense_3_utr_granges[subjectHits(over_3_utr)] |> as.tibble() |> dplyr::select(-width, -seqnames, -strand) |> dplyr::rename(start_3_utr = start, end_3_utr = end)

peak_3utr_df <- bind_cols(peak_2_3utr, ense_3utr_2_peak)
#peak_5utr_df$overlap_width <- overlap_widths_5utr
peak_3utr_df$region_class <- "3_UTR"

peak_3utr_dt<- as.data.table(peak_3utr_df)
setkey(peak_3utr_dt, canonical, region_class, external_transcript_name, rank)
peak_3utr_dt<-peak_3utr_dt[order(canonical,exon_class,external_transcript_name,rank)]
peak_3utr_unique <- peak_3utr_dt[!duplicated(id_name)]

r_anno$region_class[which(r_anno$id_name %in% peak_3utr_unique$id_name)] <- "3_UTR"
```

```{r}
rm(over_3_utr, overlap_ranges_3utr, peak_2_3utr,ense_3utr_2_peak,peak_3utr_df,peak_3utr_dt)
```

## Raw counts clustering before normalization (MDS plot)

```{r}

sample_data <- scale(t(counts_df)) ## matrix with samples on rows and regions on columns
sample_distances <- dist(sample_data, method = "euclidean") ## computes and return the distance matrix, with the distances between the rows of a data matrix

mds_cmdscale <- as.data.frame(cmdscale(as.matrix(sample_distances)))
colnames(mds_cmdscale) <- c("MDS_1", "MDS_2")
mds_cmdscale$sample <- rownames(mds_cmdscale)

mds_cmdscale <- left_join(mds_cmdscale, c_anno)

bs <- 11
bf <- "Arial"

cmd <- ggplot(mds_cmdscale, aes(MDS_1, MDS_2, colour = condition, shape = library)) + #mds plot
  geom_point(size = 2.5, alpha = .9) +
  geom_text_repel(aes(label = paste(library, cell, replicate, sep = "_")), size = 0.3*bs, fontface = "bold", show.legend = F) +
  theme_bw(base_size = bs, base_family = bf) + 
  theme(legend.position = "right", panel.grid.minor = element_blank()) +
  scale_color_manual(values = c("#5b5b5b","#bf9000","#0b5394","#cc0000"))

cmd

#ggsave(cmd, file = "./results/counts_cmdscale.pdf", device = cairo_pdf, height = 6, width = 8)
```

## Number of reads for each sample

```{r}

bar_df <- data.frame("sample" = colnames(counts_df), "count" =(colSums(counts_df)/1000000))
bar_df <- left_join(bar_df, c_anno)

bp <- ggplot(bar_df, aes(sample, count, fill = condition)) +
  geom_bar(stat="identity", width = 0.7) +
  coord_flip() +
  theme_bw(base_size = bs, base_family = bf) +
  theme(axis.title.y = element_blank(), panel.border = element_blank(), axis.text.y = element_text(size = 8, colour = "black")) +
  theme(panel.grid.major.y = element_blank(), panel.grid.minor.x = element_blank()) +
  theme(legend.position = "na")+
  labs(y = "aligned reads (mln)") +
  scale_fill_manual(values =c("#5b5b5b","#bf9000","#0b5394","#cc0000"))
  
bp  

#ggsave(bp, file = "./results/bp_librarysize.pdf", device = cairo_pdf, height = 6, width = 8)
```

## TMM Normalization

```{r edgeR, eval=FALSE}
edge_c <- DGEList(counts = counts_df, group = c_anno$condition, samples = c_anno, genes = r_anno) #create DGE object
edge_n <- calcNormFactors(edge_c, method = "TMM") #compute normalization factors with TMM method

norm_factors <- (edge_n$samples$lib.size*edge_n$samples$norm.factors)/mean(edge_n$samples$lib.size*edge_n$samples$norm.factors)
#librarysize*normfactor / mean(librarysize*normfactor)
names(norm_factors) <- edge_n$samples$sample

norm_df <- data.frame("sample" = names(norm_factors),
                      "div_norm" = norm_factors)

```


## FPKM

```{r fpkm_tables, messag = F, warning=F, echo=F, eval=F}
#fragments per kilobase  million = rpkm function
fpkm_table <- as.data.frame(rpkm(edge_n, gene.length = edge_n$genes$length, log = F))


# l_fpkm_table <- fpkm_table |> 
#                 mutate(id = rownames(fpkm_table)) |>
#                 pivot_longer(-id, names_to = "sample", values_to = "fpkm")
# l_fpkm_table <- left_join(l_fpkm_table, c_anno)
# 
# fpkm_df <- l_fpkm_table |>
#            dplyr::group_by(condition, id) |>
#            dplyr::summarise(N = n(), avg = mean(fpkm), sd = sd(fpkm)) |> 
#            ungroup() |>
#            mutate("se" = sd/sqrt(N))
# 
# fpkm_avg <- fpkm_df %>% dplyr::select(id,condition,avg) %>% pivot_wider(names_from = condition, values_from = avg)
# colnames(fpkm_avg)[-1]<-paste0(colnames(fpkm_avg)[-1],"_fpkm_avg")
# 
# fpkm_se <- fpkm_df %>% dplyr::select(id,condition,se) %>% pivot_wider(names_from = condition, values_from = se)
# colnames(fpkm_se)[-1]<-paste0(colnames(fpkm_se)[-1],"_fpkm_se")
# 
# fpkm_avgse<-left_join(fpkm_avg,fpkm_se)
# rownames(fpkm_avgse) <- fpkm_avgse$id

#write_tsv(fpkm_table, "~/Desktop/YBX1_clip_fpkm.tsv")
```

## PCA (based on FPKM)

```{r warning=F, message=F}
#prcomp function computes the pca on the given data matrix
data.PC <- prcomp(as.matrix(t(fpkm_table)), scale. = T)

data.PC_df <- as.data.frame(data.PC$x[,1:2])
data.PC_df$sample <- rownames(data.PC_df)
data.PC_df <- left_join(data.PC_df, c_anno)

summ <- summary(data.PC)
summ$importance[2,]

pca <- ggplot(data.PC_df, aes(PC1, PC2, colour = condition, shape = library)) + 
  geom_point(size = 2.5, alpha = .9) +
  geom_text_repel(aes(label = paste(library, cell, replicate, sep = "_")), size = 0.2*bs, fontface = "bold", show.legend = F) + #adds labels directly on the plot
  scale_colour_manual(values = c("#5b5b5b","#bf9000","#0b5394","#cc0000")) +
  theme_bw(base_size = bs, base_family = bf) +
  ylab(paste0("PC2 (", round(summ$importance[2,2]*100), " %)")) +
  xlab(paste0("PC1 (", round(summ$importance[2,1]*100), "%)")) +
  theme(legend.position = "right", panel.grid.minor = element_blank())

pca

#ggsave(pca, file = "./results/fpkm_pca.pdf", height = 6, width = 8)
ggsave(pca, file = "/home/giobc/Desktop/eCLIP_final_figures/pca_fpkm.pdf", device = cairo_pdf, height = 3, width = 4)

#write_tsv(data.PC_df, "~/Desktop/YBX1_clip_PCA_values.tsv")
```

```{r standard deviation plot, fig.width=10}

sdev <- as.data.frame(data.PC$sdev)
colnames(sdev) <- "sd"
sdev$PC <- paste("PC", rownames(sdev), sep = "")

var_explained <- (sdev$sd)^2/sum((sdev$sd)^2)
sdev$var_expl <- var_explained
levels(sdev$PC) <- unique(sdev$PC)

sdev_plot <- ggplot(sdev, aes(PC, sd)) +
  geom_bar(stat = "identity", width = 0.7, fill ="#43bccd", color = "#43bccd") +
  theme_bw(base_size = bs, base_family = bf) +
  theme(axis.title.y=element_blank(),panel.border=element_blank(),axis.text.y = element_text(size=8, colour = "black")) +
  theme(panel.grid.major.y = element_blank(),panel.grid.minor.x = element_blank())+
  theme(legend.position = "na") + 
  labs(title = "Standard deviation plot", x = NULL, y = " (%)") +
  scale_x_discrete(limits=sdev$PC)

# percentage of variance explained
scree_plot <- fviz_screeplot(data.PC)

sdev_plot + scree_plot
```

## MDS plot based on FPKM values

```{r mds_fpkm}

sample_data <- scale(t(fpkm_table))
sample_distances <- dist(sample_data, method = "euclidean")

mds_cmdscale <- as.data.frame(cmdscale(as.matrix(sample_distances)))
colnames(mds_cmdscale) <- c("MDS_1", "MDS_2")
mds_cmdscale$sample <- rownames(mds_cmdscale)

mds_cmdscale <- left_join(mds_cmdscale, c_anno)

cmd <- ggplot(mds_cmdscale, aes(MDS_1, MDS_2, colour = condition, shape = library)) +
  geom_point(size = 2.5, alpha =.9) +
  geom_text_repel(aes(label = paste(library, cell, replicate, sep = "_")), size = 0.2*bs, fontface = "bold", show.legend = F) +
  scale_color_manual(values = c("#5b5b5b","#bf9000","#0b5394","#cc0000")) +
  theme_bw(base_size = bs, base_family = bf) +
  theme(legend.position = "right", panel.grid.minor = element_blank())

cmd

#ggsave(cmd, file = "results/fpkm_cmdscale.pdf", device = cairo_pdf,height = 6, width = 8)
```

## Differentially bound regions (DE peaks)

```{r design}

design <- model.matrix(~0+condition, data = edge_c$samples) # create a binary matrix to pair replicates (description of the conditions)
colnames(design) <- levels(edge_c$samples$condition)

edge_d <- estimateDisp(edge_n, design, robust = T) # robust = TRUE to have a better estimation of the outliers

edge_f <- glmQLFit(edge_d, design, robust = T) #squeeze of the variances -> estimation of the quasi-likelihood (QL) dispersions around

design
```

```{r}
edgeRglmQLF_2022 <- function(glmfit, # object of class DGEGLM
                      contrast_chr, # comparison (string)
                      comp_name="", # include in label column
                      signal_col="CPM", # signal column
                      signal_thr=0.1, # signal threshold
                      fc_col="log2_FC",
                      fc_thr=0.5, # fold change threshold
                      pval_col="p_val", # p-value column
                      pval_thr=0.05) # p-value threshold
                      { 
  
  contro<-makeContrasts(contrasts=contrast_chr,levels=glmfit$design)
  
  degs<-glmQLFTest(glmfit,
                   contrast= contro)$table[,c("logFC","PValue")]
  colnames(degs)<-c("log2_FC","p_val")
  
  degs$id<-rownames(degs)
  degs$comp<-ifelse(comp_name!="",comp_name,contrast_chr)
  degs$p_adj<-p.adjust(degs$p_val, method ="BH")
  
  active_levels<-rownames(contro)[which(contro!=0)]
  active_samples<-which(glmfit$samples$group%in%active_levels)
  cpm_sele<-cpm(glmfit,log=F)[,active_samples]
  
  degs$CPM<-round(rowMeans(cpm_sele),3)
  #degs$CPM<-round(2^degs_u$log2_CPM,3)
  
  degs$class<-"="
  degs[which(degs[,signal_col]>=signal_thr & 
             degs[,fc_col]>=fc_thr & 
             degs[,pval_col]<=pval_thr),"class"]<-"+"

  degs[which(degs[,signal_col]>=signal_thr &
             degs[,fc_col]<=(-fc_thr) &
             degs[,pval_col]<=pval_thr),"class"]<-"-"
  
  degs<-degs %>% dplyr::select(id,comp,class,log2_FC,p_val,p_adj,CPM)

  return(degs)
}

```


```{r}
# extract results
degs_w_df <- NULL
degs_l_df <- NULL

contro_list<-list(#"IP_enr"="((IP_SCC+IP_HK)/2-(Input_SCC+Input_HK)/2)",
                  "SCC_enr"="IP_SCC-Input_SCC",
                  "HK_enr"="IP_HK-Input_HK",
                  "SCC_HK_IP"="IP_SCC-IP_HK",
                  "SCC_HK_Input"="Input_SCC-Input_HK",
                  "SCC_HK_dd"="(IP_SCC-IP_HK)-(Input_SCC-Input_HK)")

for(contro_label in names(contro_list)){
  
  degs_u <- edgeRglmQLF_2022(glmfit = edge_f,
                             contrast_chr = contro_list[[contro_label]],
                             comp_name = contro_label,
                             signal_col = "CPM", signal_thr = -Inf,
                             fc_col = "log2_FC", fc_thr = 1,
                             pval_col = "p_val", pval_thr = 0.05)
  
  degs_l_df <- bind_rows(degs_l_df, degs_u)
  rownames(degs_l_df) <- NULL
  
  degs_add <- degs_u |> dplyr::select("class", "log2_FC", "p_val", "p_adj", "CPM")
  colnames(degs_add) <- paste(degs_u[1, "comp"], colnames(degs_add), sep = "_")
  degs_w_df <- bind_cols(degs_w_df, degs_add)
  
  }

degs_w_df <- bind_cols(r_anno, degs_w_df) #whole dataframe of differentially bound regions + region annotations
degs_l_df$comp <- factor(degs_l_df$comp, levels = c( "SCC_HK_dd","SCC_HK_Input","SCC_HK_IP","SCC_enr","HK_enr")) # long dataframe of differentially bound regions

#write_tsv(degs_w_df, file = "./DEGs_w_df.tsv")
#write_tsv(degs_l_df, file = "./DEGs_l_df.tsv")
```

```{r summary of the number of differentially bound peaks}
#degs_l_df <- read_tsv("results/DEGs_l_df.tsv")
#degs_w_df <- read_tsv("results/DEGs_w_df.tsv")

degs_l_df |> dplyr::group_by(comp, class) |> summarise("number" = n()) |> filter(class != "=")
```

```{r plotting function}
plot_DEGs<-function(degs_df, # dataframe containing DEGs result
                    contrast_col="comp", # Name of the column with contrast names
                    class_col="class", # Name of the column with classes (+,-,= are expected)
                    color_vec=c("grey35","grey65") # Colors for Up and Down regulated
                     ){
  res_mat<-degs_df %>% group_by(get(contrast_col),get(class_col),.drop=FALSE) %>% summarise("number"=n()) %>% ungroup() 
  # get() to return the value of variable 
  # in this way we can assign the name of the column to a variable and keep the same script changing every time only the name of the column
  # .drop=FALSE to keep the comparisons with zero-value classes (for example in the fdr df)
  colnames(res_mat)<-c("contrast","class","number")
  res_df<-res_mat %>% filter(class!="=")
  
  res_df$class<-factor(res_df$class,levels=c("+","-"))
  res_df$contrast<-as.factor(res_df$contrast)
  res_df$number<-as.numeric(res_df$number)
  res_df$number[which(res_df$class=="-")]<-res_df$number[which(res_df$class=="-")]*(-1) 
  # *(-1) to assign to the total number of down-reg a - sign and put them on the left of the plot
  
  bs<-11
  bap<-ggplot(res_df,aes(x=contrast,y=number,fill=class)) +
    scale_fill_manual(name="",values=color_vec,labels=c("Up", "Down"))+
    geom_bar(stat="identity",position=position_nudge(),width=0.7,alpha=0.9) +
    geom_text(aes(x=contrast, y=number, label=abs(number), hjust=ifelse(number>=0,1.1,-0.1)), colour="white",size=bs*0.3,fontface="plain") +
    theme_tufte(base_size = bs,base_family="Arial") +
    theme(axis.title.y=element_blank()) +
    theme(panel.border=element_blank()) +
    theme(axis.ticks=element_blank()) +
    theme(panel.grid=element_blank()) +
    geom_hline(yintercept=0,col="grey30",lty=1,lwd=0.03*bs)+
    theme(axis.text.y=element_text(colour="grey20")) +
    theme(axis.text.x=element_blank()) +
    theme(legend.position="none") +
    #theme(legend.justification=c(1.05,1.05), legend.position=c(1,0.07)) +
    theme(legend.title=element_blank())+
    coord_flip()+
    annotate("text", x=Inf, y=0, label = "Up",hjust=(-0.3),vjust=(1),colour=color_vec[1])+
    annotate("text", x=Inf, y=0, label = "Down",hjust=(1.3),vjust=(1),colour=color_vec[2])+
    theme(legend.text = element_text(size = 0.8*bs),
          legend.key.size = unit((0.012*bs),"in"))+
    labs(y="Number of DEGs")
  
  return(bap)
}
```


```{r plot of DE peaks}

in_bp <- degs_l_df |> filter((comp == "SCC_HK_IP" | comp == "SCC_enr" | comp == "HK_enr"))

in_bp <- in_bp[-which(in_bp$comp == "SCC_enr" & in_bp$class == "-"),]
in_bp <- in_bp[-which(in_bp$comp == "HK_enr" & in_bp$class == "-"),]

pDEGs<-plot_DEGs(in_bp,
          contrast_col="comp", 
          color_vec=c("#cc0000","#0b5394"),
          class_col="class")

#"#F21A00","#f7b32b", "#046C9A", "#273046"

pDEGs

#ggsave(pDEGs, file = "results/DE_regions.pdf", device = cairo_pdf, height = 3, width = 4)
```

In the plot we have the number of differentially bound regions. The peaks enriched in the HK are 31948 while those enriched in the SCC are 33399.

```{r}
degs_cand <- degs_w_df |> dplyr::filter((SCC_enr_class == "+" | HK_enr_class == "+")) # filter to have in the df only SCC or HK enriched peaks

fc_col <- "SCC_HK_IP_log2_FC" 
pval_col <- "SCC_HK_IP_p_val"

fc_thr <- 0.75
pval_thr <- 0.05

degs_cand$eclip_class <- "common"
degs_cand[which(degs_cand[,fc_col] >= fc_thr & degs_cand[,pval_col]<= pval_thr), "eclip_class"] <- "SCC_gain" 
degs_cand[which(degs_cand[,fc_col] <= (-fc_thr) & degs_cand[,pval_col]<= pval_thr), "eclip_class"] <- "HK_gain"

fc_col <- "SCC_HK_dd_log2_FC"

degs_cand$eclip_cor_class <- degs_cand$eclip_class
degs_cand[which(degs_cand[,"eclip_class"] == "SCC_gain" & degs_cand[,fc_col] >= fc_thr), "eclip_cor_class"] <- "SCC_gain_delta"
degs_cand[which(degs_cand[,"eclip_class"] == "HK_gain" & degs_cand[,fc_col] <= (-fc_thr)), "eclip_cor_class"] <- "HK_gain_delta"


table(degs_cand$eclip_class)

#write_tsv(degs_cand, file = "./DEGs_cand.tsv")
```

## Volcano plots

```{r}
in_df <- degs_cand
in_df$HK_enr_eclip_class <- ifelse(in_df$HK_enr_class == "+", TRUE, FALSE)
in_df$SCC_enr_eclip_class <- ifelse(in_df$SCC_enr_class == "+", TRUE, FALSE)

ggplot(in_df, aes(x = SCC_enr_log2_FC, y = -log(SCC_enr_p_val), color = SCC_enr_eclip_class)) +
  geom_point(alpha = 0.8) +
  scale_color_manual(values = c("grey50","#b51416")) +
  theme_bw() +
  theme(panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank()) +
  labs(color = "Enriched") 

#ggsave("./results/SCC_enr_volcano.png", device = "png", width = 6, height = 5)

ggplot(in_df, aes(x = HK_enr_log2_FC, y = -log(HK_enr_p_val), color = HK_enr_eclip_class)) +
  geom_point(alpha = 0.8) +
  scale_color_manual(values = c("grey50","#240ba2")) +
  theme_bw() +
  theme(panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank()) +
  labs(color = "Enriched") 

#ggsave("./results/HK_enr_volcano.png", device = "png", width = 6, height = 5)
```

## Number of peaks

```{r number of peaks plot}
peaks_plot_df <- degs_cand |> group_by(eclip_cor_class) |> summarise(peaks = n()) |> ungroup()

peaks_plot <- ggplot(peaks_plot_df) +
  theme_bw(base_family = "Arial", base_size = bs) +
  #facet_wrap(vars(region_class), ncol = 1) +
  xlab("") + ylab("Peaks (N)") +
  #ggtitle("Peaks numbers") +
  theme(plot.title = element_text(hjust = 0.5)) + guides(fill = guide_legend(title = "class")) +
  theme(axis.title.y = element_blank(), panel.border = element_blank()) +
  theme(panel.grid.major.y = element_blank(), panel.grid.minor.x = element_blank())+
  geom_col(aes(x= eclip_cor_class, y = peaks, fill = eclip_cor_class), width = .5, alpha = .9, position = "dodge") + 
  scale_fill_manual(values=c("grey65","#6c56da","#240ba2","#eb5e60","#b51416")) +
  geom_text_repel(aes( x = eclip_cor_class, y = peaks, label = peaks, vjust = 0.5, hjust = 0, family = "bf"), segment.color = NA, color = "white") +
  theme(legend.position = "none") +
  coord_flip()

peaks_plot

#ggsave(peaks_plot, file = "results/peaks_number_class.pdf", device = cairo_pdf, height = 6, width = 8)
#ggsave("~/Desktop/CLIP_peaks_class_numbers.pdf", device = cairo_pdf, height = 3, width = 4)

```

## Region analysis

```{r}
region_df <- degs_cand |> 
  filter(eclip_class != "common") |>
  group_by(region_class, eclip_class) |> 
  summarise(peaks = n()) |> 
  ungroup() |> 
  group_by(eclip_class) |>
  mutate(tot_peaks = sum(peaks)) |>
  ungroup() |>
  mutate(peak_perc = round((peaks/tot_peaks)*100, 2))

region_df$region_class <- factor(region_df$region_class, levels = c("3_UTR", "CDS", "5_UTR", "Exon", "Intron", "Intergenic"))

ggplot(region_df, aes(x = region_class, y = peak_perc, fill = eclip_class)) + 
  geom_bar(stat = "identity") + 
  theme_tufte(base_size = bs, base_family = bf) + 
  facet_wrap(~eclip_class) + 
  coord_flip() +
  scale_fill_manual(values = rev(c("#5e548e","#6b9080"))) +
  theme(legend.position = "none", axis.title.y = element_blank()) +
  labs(y = "% peaks") +
  #labs(y = "Number m5Cs") +
  geom_text(aes(label = paste(round(peak_perc,1), "%"), hjust = ifelse(region_df$region_class %in% c("3_UTR", "CDS"),1,-0.1)))

#write.xlsx(region_df |> dplyr::select(region_class,eclip_class,peak_perc), file = "~/Desktop/Projects/YBX1_giorgia/eClip/peaks_localization.xlsx")
```

## Scatterplot

```{r}
scatter_enr_df <- degs_w_df
scatter_enr_df$overall_enr_class <- FALSE
scatter_enr_df$overall_enr_class[which(scatter_enr_df$SCC_enr_class == "+")] <- TRUE
scatter_enr_df$overall_enr_class[which(scatter_enr_df$HK_enr_class == "+")] <- TRUE

in_bp <- scatter_enr_df |> dplyr::select(HK_enr_log2_FC, SCC_enr_log2_FC, overall_enr_class)

ggplot(scatter_enr_df, aes(x = HK_enr_log2_FC, y = SCC_enr_log2_FC, color = overall_enr_class, order = as.numeric(overall_enr_class))) +
  geom_point() +
  theme_bw() +
  scale_colour_manual(values=rev(c("#a6a6a6","grey30"))) +
  theme(panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank()) +
  labs(color = "Enriched")

#ggsave("~/Desktop/eCLIP_final_figures/scatter_enriched_peaks.png", device = "png", width = 5, height = 4)
```


```{r}

scatter_plot <- ggplot(degs_cand, aes(SCC_HK_Input_log2_FC, SCC_HK_IP_log2_FC, colour = eclip_cor_class)) +
  geom_point(size = 1, alpha =.5, shape = 16) +
  theme_bw(base_size = bs, base_family = bf) +
  geom_vline(xintercept = 0, color = "grey50", lty = 2) +
  geom_hline(yintercept = 0, color = "grey50", lty = 2) +
  scale_colour_manual(values=c("grey65","#6c56da","#240ba2","#eb5e60","#b51416"))

scatter_plot

#ggsave(scatter_plot, file = "results/scatter_v1.pdf", device = cairo_pdf, height = 6, width = 8)
#ggsave("~/Desktop/CLIP_scatter.pdf", device = cairo_pdf, height = 3, width = 4)
```