---
title: "YBX1_Nanopore_DRS"
output: html_document
date: "2025-09-12"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Dorado V1

```{r}
library(tidyverse)
library(data.table)
library(xlsx2dfs)
library(GenomicRanges)
library(BSgenome.Hsapiens.UCSC.hg38)
library(ggpubr)
library(ggbeeswarm)
library(ggthemes)
library(ggrepel)
library(ggbreak)
library(rstatix)
library(ggstatsplot)


library(extrafont) # plots with arial font (and cairo_pdf device)
loadfonts()
bs <- 11
bf <- "Arial"

colors_list <- c("common" = "#a6a6a6", "SCC" = "#ae68ff", "HK" = "#40ab7c")
colors <- c("#a6a6a6","#ae68ff","#40ab7c")
```

```{r}
## Data Loading
#dir <- "/data/giorgia/YBX1/Nanopore/Dorado_v1_SUP_pileup/filtered/merged/"

read_modkit_methyltable <- function(dir){
  
  list_input <- fs::dir_ls(dir,glob="*.bed", type = "file")
  
  for (i in (list_input)){
    
    s_name <- str_split(i,dir,simplify = T)[,2]
    s_name <- str_split(s_name, "\\.", simplify=T)[,1]
    s_name <- str_split(s_name, "_", simplify=T)[,2]
    #s_name <- rlang::sym(s_name)
    
    print(paste("Processing", s_name))
    
    i_tab <- read_tsv(i, col_names = F, progress = F, show_col_types = F)
    i_tab <- i_tab[,-c(7,8,9)]
    colnames(i_tab) <- c("chromosome", "start", "end", 
                         "modification", "score", "strand", 
                         "N_valid_cov", "percent_modified", "N_mod", 
                         "N_canonical", "N_other_mod", "N_delete", 
                         "N_fail", "N_diff", "N_nocall")
    
    i_tab$id <- paste(i_tab$chromosome, i_tab$start, i_tab$end, 
                      i_tab$strand ,i_tab$modification, sep = "_")
    
    i_tab_cov <- i_tab |>
      dplyr::select(id, N_valid_cov)
    colnames(i_tab_cov)[2] <- s_name
    
    i_tab_perc <- i_tab |>
      dplyr::select(id, percent_modified)
    colnames(i_tab_perc)[2] <- s_name

    i_tab_mod <- i_tab |>
      dplyr::select(id, N_mod)
    colnames(i_tab_mod)[2] <- s_name
    
    if(!exists("cov_df")){
      cov_df <- i_tab_cov
      mod_df <- i_tab_mod
      perc_df <- i_tab_perc
    } else {
      cov_df <- full_join(cov_df, i_tab_cov, by = "id")
      mod_df <- full_join(mod_df, i_tab_mod, by = "id")
      perc_df <- full_join(perc_df, i_tab_perc, by = "id")
    }
  }
  
  return(list("perc_df" = perc_df, "cov_df" = cov_df, "mod_df" = mod_df))
}
```

```{r}
Dorado_v1_tables <- read_modkit_methyltable(dir = dir)

mod_df <- Dorado_v1_tables$mod_df |> replace_na(replace = list(Sample_1 = 0, Sample_2 = 0, Sample_3 = 0,
                                                               Sample_4 = 0, Sample_5 = 0, Sample_6 = 0))
cov_df <- Dorado_v1_tables$cov_df |> replace_na(replace = list(Sample_1 = 0, Sample_2 = 0, Sample_3 = 0,
                                                               Sample_4 = 0, Sample_5 = 0, Sample_6 = 0))
perc_df <- Dorado_v1_tables$perc_df |> replace_na(replace = list(Sample_1 = 0, Sample_2 = 0, Sample_3 = 0,
                                                               Sample_4 = 0, Sample_5 = 0, Sample_6 = 0))
```

```{r}
c_anno <- data.frame("sample" = colnames(mod_df)[-1])
c_anno$condition <- c(rep("HK",3), rep("SCC", 3))
#c_anno$N_type <- str_split_i(c_anno$sample, "_", i = 3)
```

```{r}
r_anno <- data_frame("id" = mod_df$id)

r_anno_add <- as.data.frame(str_split(r_anno$id, "_", simplify = T))
colnames(r_anno_add) <- c("chr", "start", "end", "strand", "modification")

r_anno <- cbind(r_anno, r_anno_add)
rownames(r_anno) <- r_anno$id

r_anno$modification <- as.character(r_anno$modification)
r_anno$modification[which(r_anno$modification == "m")] <- "m5C"
r_anno$modification[which(r_anno$modification == "a")] <- "m6A"
r_anno$modification[which(r_anno$modification == "17802")] <- "psi"
r_anno$modification[which(r_anno$modification == "17596")] <- "inosine"
r_anno$modification[which(r_anno$modification == "19227")] <- "2OmeU"
r_anno$modification[which(r_anno$modification == "19228")] <- "2OmeC"
r_anno$modification[which(r_anno$modification == "69426")] <- "2OmeA"
r_anno$modification <- factor(r_anno$modification, levels = unique(r_anno$modification))

rownames(r_anno) <- r_anno$id
```

### Filtering

#### Coverage filtering

We apply a preliminary filter on the coverage, to retain only modified sites with a coverage >= 10 in at least two replicates of a condition.

```{r}
cov_dt <- as.data.table(cov_df)
c_anno_dt <- as.data.table(c_anno)

cov_l_dt <- cov_dt[,id:=as.factor(id)]
cov_l_dt <- melt.data.table(cov_l_dt, id.vars = c("id"), variable.name = "sample", value.name = "counts")
cov_l_dt <- cov_l_dt[c_anno, on = "sample"]

cov_filter_dt <- cov_l_dt[,.(min_c = sum(counts >= 10)), .(id, condition)][condition %in% c("HK", "SCC")]
cov_filter_dt <- cov_filter_dt[,.(passes_c = sum(min_c >= 2)), .(id)] 
cov_filter_id <- as.character(cov_filter_dt[passes_c > 0]$id) # 1.898.038 bases retained
```

1.898.038 pass the coverage filter

```{r}
## filtering the modification dataframe to retain only the sites passing the coverage threshold
rownames(mod_df) <- mod_df$id
mod_cov_filt <- mod_df[cov_filter_id,] ## 1.898.038 modified bases

#save(cov_df, cov_filter_id, mod_cov_filt, file ="./Results/cov_filtered.RData")
```

#### Modification Filtering 

Now we perform an additional filter to retain only sites with at least one base supporting the specific modification in at least two different samples.

```{r}
mod_md_dt <- as.data.table(mod_cov_filt)

sig_thr <- 2 #count threshold

mod_long_dt <- mod_md_dt[,id:=as.factor(id)]
mod_long_dt <- melt.data.table(mod_long_dt, id.vars = c("id"), variable.name = "sample", value.name = "counts")
 
setkey(mod_long_dt, sample) #sorts a data.table and marks it as sorted, the sorted columns are the key, always in ascending order
setkey(c_anno_dt, sample)
 
mod_long_dt <- mod_long_dt[c_anno_dt, on = "sample"]

mod_filter_dt <- mod_long_dt[,.(min_c = sum(counts > 0)), .(id, condition)][condition %in% c("HK", "SCC")]
mod_filter_dt <- mod_filter_dt[,.(passes_c = sum(min_c >= sig_thr)), .(id)] 
md_filter_id <- as.character(mod_filter_dt[passes_c > 0]$id) ## 1.898.038
```

#### Filtered dfs

```{r}
filt_mod_df <- mod_df[which(mod_df$id %in% md_filter_id),]
r_anno <- r_anno[which(r_anno$id %in% md_filter_id),]

# Convert percent_mod to proportions
perc_df <- perc_df[which(perc_df$id %in% md_filter_id),]
prop_data <- perc_df[,-1]/100
```

## Annotation

```{r}
#load("~/rdds_shared/data/giorgia/YBX1/Nanopore/Analysis/Dorado_V1/Dorado_v1_SUP_filtered_modification.RData")
load("~/Desktop/anno/ens_anno_110_hs.RData")
```


```{r}
# creating exon and gene annotation
ense_slim <- ense |> dplyr::select(ensembl_exon_id,
                                   rank, 
                                   exon_chrom_start,
                                   '5_utr_start',
                                   "5_utr_end",
                                   genomic_coding_start,
                                   genomic_coding_end,
                                   "3_utr_start",
                                   "3_utr_end",
                                   exon_chrom_end,
                                   l_exon,
                                   l_5utr,
                                   l_3utr,
                                   l_cds,
                                   TIS, TTS, exon_class,
                                   ensembl_transcript_id) 


# transcript annotation
enst_slim <- enst |> dplyr::select(external_transcript_name,
                                   transcript_biotype, transcript_length,
                                   canonical, ensembl_transcript_id, ensembl_gene_id)

## We filter to retain only protein coding transcripts for protein coding genes
# to avoid annotation override based on other biotypes
enst_filt <- enst_slim |> 
  group_by(ensembl_gene_id, transcript_biotype) |> 
  summarise(N = n(), transcripts = paste(external_transcript_name, collapse = " "))
enst_filt$coding <- ifelse(enst_filt$transcript_biotype == "protein_coding", T, F)

coding_ense <- enst_filt |> filter(coding == T) |> pull(ensembl_gene_id) |> unique()
enst_filt_n_cod <- enst_filt |> dplyr::filter(ensembl_gene_id %in% coding_ense & transcript_biotype != "protein_coding")
excluded_t <- unlist(strsplit(enst_filt_n_cod$transcripts, " "))

enst_slim_cod <- enst_slim |> dplyr::filter(!external_transcript_name %in% excluded_t)
ense_slim_cod <- ense_slim |> dplyr::filter(ensembl_transcript_id %in% enst_slim_cod$ensembl_transcript_id) 

rownames(enst_slim) <- enst_slim$ensembl_transcript_id
enst_slim <- enst_slim[ense_slim$ensembl_transcript_id,]
enst_slim$ensembl_transcript_id <- NULL

# gene annotation
ensg_slim <- ensg |> dplyr::select(external_gene_name, ensembl_gene_id, gene_biotype, chr_ucsc, strand_ucsc)
rownames(ensg_slim) <- ensg_slim$ensembl_gene_id
ensg_slim <- ensg_slim[enst_slim$ensembl_gene_id,]
ensg_slim$ensembl_gene_id <- NULL

# single annotation 
ense_df <- left_join(enst_slim_cod, ensg, by = "ensembl_gene_id")
ense_df <- left_join(ense_df, ense_slim_cod, by = "ensembl_transcript_id")

#ense_df <- cbind(ense_slim_cod, ensg_slim) #combines all the annotation dataframes into one
# in rownames we have the gene names
#rownames(ense_df)<-make.unique(str_c(ense_df$external_transcript_name, ense_df$rank, sep="_")) #combines external transcript name and rank with "_"

# the following function takes a dataframe-like object as input and tries to automatically find the columns that describe genomic ranges
# returns a GRanges object
ense_granges <- makeGRangesFromDataFrame(ense_df,
                                         keep.extra.columns = T, 
                                         ignore.strand = F,
                                         seqinfo = NULL,
                                         seqnames.field = "chr_ucsc", 
                                         start.field = "exon_chrom_start",  
                                         end.field = "exon_chrom_end", 
                                         strand.field = "strand_ucsc",
                                         starts.in.df.are.0based = F)

ensg_df <- ensg |> dplyr::select(external_gene_name, ensembl_gene_id, gene_biotype, chr_ucsc, strand_ucsc, start_position, end_position, description)

ensg_granges <- makeGRangesFromDataFrame(ensg_df,
                                         keep.extra.columns = T, 
                                         ignore.strand = F,
                                         seqinfo = NULL,
                                         seqnames.field = "chr_ucsc", 
                                         start.field = "start_position",  
                                         end.field = "end_position", 
                                         strand.field = "strand_ucsc",
                                         starts.in.df.are.0based = F)

```

```{r}
## 5' UTR granges
ense_5_utr_granges <- makeGRangesFromDataFrame(ense_df |> dplyr::filter(!is.na(l_5utr)),
                                               keep.extra.columns = T,
                                               ignore.strand = F,
                                               seqinfo = NULL,
                                               seqnames.field = "chr_ucsc",
                                               start.field = "5_utr_start",
                                               end.field = "5_utr_end",
                                               strand.field = "strand_ucsc",
                                               starts.in.df.are.0based = F)



## CDS granges
ense_cds_granges <- makeGRangesFromDataFrame(ense_df |> dplyr::filter(!is.na(l_cds)),
                                               keep.extra.columns = T,
                                               ignore.strand = F,
                                               seqinfo = NULL,
                                               seqnames.field = "chr_ucsc",
                                               start.field = "genomic_coding_start",
                                               end.field = "genomic_coding_end",
                                               strand.field = "strand_ucsc",
                                               starts.in.df.are.0based = F)

## 3' UTR granges
ense_3_utr_granges <- makeGRangesFromDataFrame(ense_df |> dplyr::filter(!is.na(l_3utr)),
                                               keep.extra.columns = T,
                                               ignore.strand = F,
                                               seqinfo = NULL,
                                               seqnames.field = "chr_ucsc",
                                               start.field = "3_utr_start",
                                               end.field = "3_utr_end",
                                               strand.field = "strand_ucsc",
                                               starts.in.df.are.0based = F)

```

```{r}
r_anno_complete <- r_anno
r_anno <- r_anno_complete |> dplyr::select(id,chr,start,end,strand,modification)
```


```{r}
r_anno$start <- as.numeric(r_anno$start)
r_anno$end <- as.numeric(r_anno$end)

chr_conversion_canonical <- data.frame(
   "chr_ens"=c("1","2","3","4","5","6","7","8","9","10",
      "11","12","13","14","15","16","17","18","19","20",
      "21","22","X","Y","MT"),
   "chr_ucsc"=c("chr1","chr2","chr3","chr4","chr5","chr6","chr7","chr8","chr9","chr10",
      "chr11","chr12","chr13","chr14","chr15","chr16","chr17","chr18","chr19","chr20",
      "chr21","chr22","chrX","chrY","chrM"))

r_anno <- r_anno |> dplyr::filter(chr %in% chr_conversion_canonical$chr_ucsc) #keep only the canonical chromosomes

# check for peak width
mod_base_granges <- makeGRangesFromDataFrame(r_anno,
                                         keep.extra.columns = F,
                                         ignore.strand = F,
                                         seqinfo = NULL,
                                         seqnames.field = "chr",
                                         start.field = "start",
                                         end.field = "end",
                                         strand.field = "strand",
                                         starts.in.df.are.0based = T)

r_anno$peak_seq <- getSeq(BSgenome.Hsapiens.UCSC.hg38, mod_base_granges) |> as.character() |> stringr::str_replace_all("T", "U")

r_anno <- r_anno[!(str_detect(r_anno$peak_seq, "N")),] #remove sequences with N

mod_base_granges <- makeGRangesFromDataFrame(r_anno,
                                         keep.extra.columns = T,
                                         ignore.strand = F,
                                         seqinfo = NULL,
                                         seqnames.field = "chr",
                                         start.field = "start",
                                         end.field = "end",
                                         strand.field = "strand",
                                         starts.in.df.are.0based = T)
```

### genes annotation

```{r}
## Annotation on genes to find intergenic regions and distinguish them from those spanning genes

over_genes<- GenomicRanges::findOverlaps(mod_base_granges, ensg_granges) #this function returns the overlap (n.rows) between peaks and genes

### Additional ----
## indentifica le regioni di sovrapposizione
genes_overlaps_ranges <- pintersect(mod_base_granges[queryHits(over_genes)], ensg_granges[subjectHits(over_genes)])

## calcola la larghezza delle sovrapposizioni
#genes_overlap_widths <- width(genes_overlaps_ranges)


mod_base_2_ensg <- mod_base_granges[queryHits(over_genes)] %>% as_tibble() %>% 
  dplyr::select(-width)
ensg_2_mod_base<- ensg_granges[subjectHits(over_genes)] %>% as_tibble() %>%
  dplyr::rename(gene_chrom_start=start,
                gene_chrom_end=end,
                gene_width=width) %>% 
  dplyr::select(-seqnames,-strand)

mod_ensg_df <-bind_cols(mod_base_2_ensg,ensg_2_mod_base) 
#mod_ensg_df$overlap_width <- genes_overlap_widths
#peak_ensg_df <- peak_ensg_df %>% dplyr::filter(!(id_name %in% peak_ense_add$id_name))

mod_ensg_df$region_class <- "Intron"
mod_ensg_df$gene_biotype<-factor(mod_ensg_df$gene_biotype)
mod_ensg_df$gene_biotype<-relevel(mod_ensg_df$gene_biotype, ref = "protein_coding")

mod_ensg_dt<- as.data.table(mod_ensg_df)
setkey(mod_ensg_dt,gene_biotype,gene_width,external_gene_name)
mod_ensg_dt<-mod_ensg_dt[order(gene_biotype,gene_width,external_gene_name)]
mod_ensg_unique <- mod_ensg_dt[!duplicated(id)]
```

```{r}
r_anno$region_class <- "Intergenic"
r_anno$region_class[which(r_anno$id %in% mod_ensg_unique$id)] <- "Intron"
```

```{r}
rm(over_genes,genes_overlaps_ranges,mod_base_2_ensg,ensg_2_mod_base, mod_ensg_df,mod_ensg_dt)
gc()
```

```{r}
## Annotation of peaks over exonic regions (still no details about 5', CDS, 3')
over_exons <- GenomicRanges::findOverlaps(mod_base_granges, ense_granges)

## indentifica le regioni di sovrapposizione
exons_overlaps_ranges <- pintersect(mod_base_granges[queryHits(over_exons)], ense_granges[subjectHits(over_exons)])

## calcola la larghezza delle sovrapposizioni
#exons_overlap_widths <- width(exons_overlaps_ranges)

mod_base_2_ense <- mod_base_granges[queryHits(over_exons)] %>% as_tibble() %>% 
  dplyr::select(-width)
ense_2_mod_base<- ense_granges[subjectHits(over_exons)] %>% as_tibble() %>%
  dplyr::rename(exon_chrom_start=start,
                exon_chrom_end=end,
                exon_width=width) %>% 
  dplyr::select(-seqnames,-strand)

mod_ense_df <-bind_cols(mod_base_2_ense,ense_2_mod_base) 
#mod_ense_df$overlap_width <- exons_overlap_widths
#peak_ensg_df <- peak_ensg_df %>% dplyr::filter(!(id_name %in% peak_ense_add$id_name))

mod_ense_df$region_class <- "Exon"
mod_ense_df$gene_biotype<-factor(mod_ense_df$gene_biotype)
mod_ense_df$gene_biotype<-relevel(mod_ense_df$gene_biotype, ref = "protein_coding")

mod_ense_dt<- as.data.table(mod_ense_df)
setkey(mod_ense_dt,gene_biotype,exon_width,external_gene_name)
mod_ense_dt<-mod_ense_dt[order(gene_biotype,exon_width,external_gene_name)]
mod_ense_unique <- mod_ense_dt[!duplicated(id)]

r_anno$region_class[which(r_anno$id %in% mod_ense_unique$id)] <- "Exon"

## Add annotations of the exons 
# peak_anno <- peak_ensg_unique
# peak_anno$region_class[which(peak_ensg_unique$id_name %in% peak_ense_unique$id_name)] <- peak_ense_unique$region_class
# 
mod_anno_ense_eng <- full_join(mod_ense_unique, mod_ensg_unique)
mod_anno_ense_ensg_dt<- as.data.table(mod_anno_ense_eng)
setkey(mod_anno_ense_ensg_dt,gene_biotype,exon_width,external_gene_name)
mod_anno_ense_ensg_dt<-mod_anno_ense_ensg_dt[order(gene_biotype,exon_width,external_gene_name)]
mod_anno_ense_ensg_unique <- mod_anno_ense_ensg_dt[!duplicated(id)]
```

```{r}
rm(over_exons,exons_overlaps_ranges,mod_base_2_ense,ense_2_mod_base,mod_ense_df,mod_ense_dt,mod_anno_ense_eng,mod_anno_ense_ensg_dt)
gc()
```


### transcript region annotation
```{r over 5'utr}
over_5_utr <- GenomicRanges::findOverlaps(mod_base_granges, ense_5_utr_granges) ## overlap between peaks and 5'UTR

overlap_ranges_5utr <- pintersect(mod_base_granges[queryHits(over_5_utr)], ense_5_utr_granges[subjectHits(over_5_utr)])
#overlap_widths_5utr <- width(overlap_ranges_5utr)

mod_base_2_5utr <- mod_base_granges[queryHits(over_5_utr)] |> as.tibble() #|> dplyr::rename(_width = width)
ense_5utr_2_mod_base <- ense_5_utr_granges[subjectHits(over_5_utr)] |> as.tibble() |> dplyr::select(-width, -seqnames, -strand) |> dplyr::rename(start_5_utr = start, end_5_utr = end)

mod_base_5utr_df <- bind_cols(mod_base_2_5utr, ense_5utr_2_mod_base)
#mod_base_5utr_df$overlap_width <- overlap_widths_5utr
mod_base_5utr_df$region_class <- "5_UTR"

mod_base_5utr_dt<- as.data.table(mod_base_5utr_df)
setkey(mod_base_5utr_dt, canonical, region_class, external_transcript_name, rank)
mod_base_5utr_dt<-mod_base_5utr_dt[order(canonical,exon_class,external_transcript_name,rank)]
mod_base_5utr_unique <- mod_base_5utr_dt[!duplicated(id)]

r_anno$region_class[which(r_anno$id %in% mod_base_5utr_unique$id)] <- "5_UTR"
```

```{r}
rm(over_5_utr, overlap_ranges_5utr, mod_base_2_5utr,ense_5utr_2_mod_base,mod_base_5utr_df,mod_base_5utr_dt)
gc()
```

```{r}
over_cds <- GenomicRanges::findOverlaps(mod_base_granges, ense_cds_granges) ## overlap between peaks and 5'UTR

overlap_ranges_cds <- pintersect(mod_base_granges[queryHits(over_cds)], ense_cds_granges[subjectHits(over_cds)])
#overlap_widths_5utr <- width(overlap_ranges_5utr)

mod_base_2_cds <- mod_base_granges[queryHits(over_cds)] |> as.tibble() #|> dplyr::rename(_width = width)
ense_cds_2_mod_base <- ense_cds_granges[subjectHits(over_cds)] |> as.tibble() |> dplyr::select(-width, -seqnames, -strand) |> dplyr::rename(start_cds = start, end_cds = end)

mod_base_cds_df <- bind_cols(mod_base_2_cds, ense_cds_2_mod_base)
#mod_base_5utr_df$overlap_width <- overlap_widths_5utr
mod_base_cds_df$region_class <- "CDS"

mod_base_cds_dt<- as.data.table(mod_base_cds_df)
setkey(mod_base_cds_dt, canonical, region_class, external_transcript_name, rank)
mod_base_cds_dt<-mod_base_cds_dt[order(canonical,exon_class,external_transcript_name,rank)]
mod_base_cds_unique <- mod_base_cds_dt[!duplicated(id)]

r_anno$region_class[which(r_anno$id %in% mod_base_cds_unique$id)] <- "CDS"
```

```{r}
rm(over_cds, overlap_ranges_cds, mod_base_2_cds,ense_cds_2_mod_base,mod_base_cds_df,mod_base_cds_dt)
gc()
```

```{r over 3'utr}
over_3_utr <- GenomicRanges::findOverlaps(mod_base_granges, ense_3_utr_granges) ## overlap between peaks and 5'UTR

overlap_ranges_3utr <- pintersect(mod_base_granges[queryHits(over_3_utr)], ense_3_utr_granges[subjectHits(over_3_utr)])
#overlap_widths_5utr <- width(overlap_ranges_5utr)

mod_base_2_3utr <- mod_base_granges[queryHits(over_3_utr)] |> as.tibble() #|> dplyr::rename(_width = width)
ense_3utr_2_mod_base <- ense_3_utr_granges[subjectHits(over_3_utr)] |> as.tibble() |> dplyr::select(-width, -seqnames, -strand) |> dplyr::rename(start_3_utr = start, end_3_utr = end)

mod_base_3utr_df <- bind_cols(mod_base_2_3utr, ense_3utr_2_mod_base)
#mod_base_5utr_df$overlap_width <- overlap_widths_5utr
mod_base_3utr_df$region_class <- "3_UTR"

mod_base_3utr_dt<- as.data.table(mod_base_3utr_df)
setkey(mod_base_3utr_dt, canonical, region_class, external_transcript_name, rank)
mod_base_3utr_dt<-mod_base_3utr_dt[order(canonical,exon_class,external_transcript_name,rank)]
mod_base_3utr_unique <- mod_base_3utr_dt[!duplicated(id)]

r_anno$region_class[which(r_anno$id %in% mod_base_3utr_unique$id)] <- "3_UTR"
```

```{r}
rm(over_3_utr, overlap_ranges_3utr, mod_base_2_3utr,ense_3utr_2_mod_base,mod_base_3utr_df,mod_base_3utr_dt)
gc()
```

```{r}
r_anno_complete <- left_join(r_anno_complete, r_anno |> 
                               dplyr::select(id,region_class) |>
                               dplyr::rename("new_region_class" = region_class), by = "id")
```

```{r transcript localization of modifications}
in_df <- r_anno |> summarise(N = length(unique(id)), .by = c("modification", "region_class")) |> dplyr::mutate(N_tot = sum(N), .by = "modification") |> dplyr::mutate(perc = (N/N_tot)*100, .by = "modification")
#colnames(in_df) <- c("modification","region_class", "N")
in_df$modification <- factor(in_df$modification, levels = c("m6A", "m5C", "inosine", "2OmeU", "2OmeC", "psi", "2OmeA"))
in_df$region_class <- factor(in_df$region_class, levels = c("3_UTR", "CDS", "5_UTR", "Exon", "Intron", "Intergenic"))

ggplot(in_df, aes(region_class, perc, fill = modification)) +
  geom_bar(stat = "identity") + 
  facet_wrap(~modification, ncol = 3, scales = "free") +
  theme_tufte(base_family = "Arial", base_size = 11) +
  #geom_text(aes(label = N), hjust = 1) +
  #scale_fill_manual(values = c("#A4243B","#3F88C5","#FDB833")) +
  coord_flip() +
  theme(legend.position = "none") +
  theme(axis.text.x = element_text(angle = 90))

#ggsave("./Dorado_V1/Region_class_of_modification_filtered_perc.pdf", device = cairo_pdf, width = 6, height = 6)
```

## m5C analysis

```{r}
#load("./Dorado_V1/m5C/peak_seq_C/m5C_DMS.RData")
```

```{r}
m5C_id <- r_anno |> dplyr::filter(modification == "m5C" & peak_seq == "C") |> pull(id)

m5C_mod <- filt_mod_df[m5C_id,]
m5C_prop <- prop_data[m5C_id,]

m5C_r_anno <- r_anno |> filter(modification == "m5C" & peak_seq == "C") 
m5C_r_anno_clip <- m5C_r_anno |> filter(!is.na(eclip_class))
m5C_YBX1_regions_ids <- m5C_r_anno_clip |> pull(id)
```

### Normalization of m5C counts

```{r TMM normalization with EdgeR}
library(edgeR)

edge_c <- DGEList(counts = m5C_mod, group = c_anno$condition, samples = c_anno, genes = m5C_r_anno) #create DGE object
edge_n <- calcNormFactors(edge_c, method = "TMM") #compute normalization factors with TMM method

norm_factors <- (edge_n$samples$lib.size*edge_n$samples$norm.factors)/mean(edge_n$samples$lib.size*edge_n$samples$norm.factors)
#librarysize*normfactor / mean(librarysize*normfactor)
names(norm_factors) <- edge_n$samples$sample

norm_df <- data.frame("sample" = names(norm_factors),
                      "div_norm" = norm_factors)

m5C_mod_norm_df <- as.data.frame(cpm(edge_n, normalized.lib.sizes=TRUE))
```

```{r distribution of normalized m5C modifications counts}
in_l_df <- pivot_longer(m5C_mod_norm_df, cols = 1:6 ,names_to = "sample", values_to = "norm_mod_counts")

ggplot(in_l_df, aes(sample, log(norm_mod_counts))) +
  geom_violin() +
  coord_flip() +
  geom_boxplot() +
  theme_bw()

#ggsave("./Dorado_V1/m5C_norm_distribution.pdf", device = cairo_pdf, width = 8, height = 5)
```

```{r normalized MDS based on m5C modifications}
sample_data <- scale(t(m5C_mod_norm_df))
sample_distances <- dist(sample_data, method = "euclidean")

mds_cmdscale <- as.data.frame(cmdscale(as.matrix(sample_distances)))
colnames(mds_cmdscale) <- c("MDS_1", "MDS_2")
mds_cmdscale$sample <- rownames(mds_cmdscale)
#mds_cmdscale$sample[5] <- "S5_SCC"
mds_cmdscale <- left_join(mds_cmdscale, c_anno)

m5C_mds_norm <- ggplot(mds_cmdscale, aes(MDS_1, MDS_2, colour = condition)) +
  geom_point(size = 2.5, alpha = .9) + 
  geom_text_repel(aes(label = sample), size = 0.3*bs, fontface = "bold") +
  theme_bw(base_size = bs, base_family = bf) + 
  theme(legend.position = "none", panel.grid.minor = element_blank()) +
  scale_color_manual(values = c("#6b9080","#5e548e")) 
  #labs(title = "MDS plot based on uncorrected values")

m5C_mds_norm

#ggsave(plot = m5C_mds_norm, file = "./Dorado_V1/m5C_norm_mds.pdf", device= cairo_pdf, width = 5, height = 4)
```

```{r normalized PCA based on m5C modifications}
data.PC <- prcomp(as.matrix(t(m5C_mod_norm_df)), scale. = T)

data.PC_df <- as.data.frame(data.PC$x[,1:2])
data.PC_df$sample <- rownames(data.PC_df)
data.PC_df <- left_join(data.PC_df, c_anno)

summ <- summary(data.PC)
summ$importance[2,]

pca <- ggplot(data.PC_df, aes(PC1, PC2, colour = condition)) + 
  geom_point(size = 2.5, alpha = .9) +
  geom_text_repel(aes(label = sample), size = 0.3*bs, fontface = "bold", show.legend = F) + #adds labels directly on the plot
  scale_color_manual(values = c("#6b9080","#5e548e")) +
  theme_bw(base_size = bs, base_family = bf) +
  theme(legend.position = "right", panel.grid.minor = element_blank()) +
  ylab(paste("PC2 (", round(summ$importance[2,2]*100), " %)")) +
  xlab(paste("PC1 (", round(summ$importance[2,1]*100), "%)")) +
  theme(legend.position = "none")

pca

#write.xlsx(data.PC_df, file = "./Dorado_V1/m5C/peak_seq_C/pca_table.xlsx")
#ggsave(pca, file = "./Dorado_V1/m5C/peak_seq_C/m5C_pca_norm.pdf", device = cairo_pdf, width = 4, height = 3)
```

### Differential methylation analysis

```{r plotting function}
plot_DEGs<-function(degs_df, # dataframe containing DEGs result
                    contrast_col="comp", # Name of the column with contrast names
                    class_col="class", # Name of the column with classes (+,-,= are expected)
                    color_vec=c("grey35","grey65") # Colors for Up and Down regulated
                     ){
  res_mat<-degs_df %>% group_by(get(contrast_col),get(class_col),.drop=FALSE) %>% summarise("number"=n()) %>% ungroup() 
  # get() to return the value of variable 
  # in this way we can assign the name of the column to a variable and keep the same script changing every time only the name of the column
  # .drop=FALSE to keep the comparisons with zero-value classes (for example in the fdr df)
  colnames(res_mat)<-c("contrast","class","number")
  res_df<-res_mat %>% filter(class!="=")
  
  res_df$class<-factor(res_df$class,levels=c("+","-"))
  res_df$contrast<-as.factor(res_df$contrast)
  res_df$number<-as.numeric(res_df$number)
  res_df$number[which(res_df$class=="-")]<-res_df$number[which(res_df$class=="-")]*(-1) 
  # *(-1) to assign to the total number of down-reg a - sign and put them on the left of the plot
  
  bs<-11
  bap<-ggplot(res_df,aes(x=contrast,y=number,fill=class)) +
    scale_fill_manual(name="",values=color_vec,labels=c("Up", "Down"))+
    geom_bar(stat="identity",position=position_nudge(),width=0.7,alpha=0.9) +
    geom_text(aes(x=contrast, y=number, label=abs(number), hjust=ifelse(number>=0,1.1,-0.1)), colour="white",size=bs*0.3,fontface="plain") +
    theme_tufte(base_size = bs,base_family="Arial") +
    theme(axis.title.y=element_blank()) +
    theme(panel.border=element_blank()) +
    theme(axis.ticks=element_blank()) +
    theme(panel.grid=element_blank()) +
    geom_hline(yintercept=0,col="grey30",lty=1,lwd=0.03*bs)+
    theme(axis.text.y=element_text(colour="grey20")) +
    theme(axis.text.x=element_blank()) +
    theme(legend.position="none") +
    #theme(legend.justification=c(1.05,1.05), legend.position=c(1,0.07)) +
    theme(legend.title=element_blank())+
    coord_flip()+
    annotate("text", x=Inf, y=0, label = "Up",hjust=(-0.3),vjust=(1),colour=color_vec[1])+
    annotate("text", x=Inf, y=0, label = "Down",hjust=(1.3),vjust=(1),colour=color_vec[2])+
    theme(legend.text = element_text(size = 0.8*bs),
          legend.key.size = unit((0.012*bs),"in"))+
    labs(y="Number of DMSs")
  
  return(bap)
}
```

#### Proportions of m5C modifications

```{r limma DMS}
#m5C_id <- r_anno |> filter(modification == "m5C") |> pull(id)
#m5C_prop_data <- prop_data[m5C_id,]

m5C_prop_matrix <- as.matrix(m5C_prop)

design <- model.matrix(~0+c_anno$condition)
colnames(design) <- unique(c_anno$condition)
rownames(design) <- c_anno$sample

design

fit1 <- limma::lmFit(m5C_prop_matrix, design)

contro_list <- list("SCC" = "SCC-HK")
contrasts <- limma::makeContrasts(contrasts = contro_list, levels = design)
colnames(contrasts) <- names(contro_list)

fit2 <- limma::eBayes(contrasts.fit(fit1, contrasts = contrasts))
```

```{r extraction of results}
LimmaDEP_v2 <- function(fit,
                        design,
                        contrast_chr,
                        comp_name = "",
                        coef,
                        fc_col = "logFC",
                        fc_thr = 0.5,
                        pval_col = "p_val",
                        pval_thr = 0.05)
  {
  contrast_result <- topTable(fit, number = nrow(fit), coef = coef, )
  contrast_result$comp <- ifelse(comp_name != "", comp_name, contrast_chr)
  contrast_result <- contrast_result |> dplyr::rename(p_val = "P.Value" , adj_pval= "adj.P.Val") 
  contrast_result$id <- rownames(contrast_result)
  contrast_result <- contrast_result |> arrange(id)
  
  contrast_result$class <- "="
  contrast_result[which(contrast_result[,fc_col] >= fc_thr &
                          contrast_result[,pval_col] <= pval_thr), "class"] <- "+"
  contrast_result[which(contrast_result[,fc_col] <= (-fc_thr) &
                          contrast_result[,pval_col] <= pval_thr), "class"] <- "-"
  
  return(contrast_result)
  
}

m5C_prop_dms_l_df <- NULL
m5C_prop_dms_df <- data.frame("id" = rownames(m5C_prop_matrix))
#m5C_prop_dms_df <- data_frame()

m5C_prop_dms_l_df <- LimmaDEP_v2(fit = fit2,
                                 contrast_chr = contro_list$SCC,
                                 comp_name = names(contro_list),
                                 coef = which(colnames(fit2$coefficients) == names(contro_list)),
                                 fc_thr = 0.02, pval_thr = 0.05)


dms_add <- m5C_prop_dms_l_df |> dplyr::select(logFC, AveExpr, p_val, adj_pval, comp, class) # add class when fixed
colnames(dms_add) <- paste(m5C_prop_dms_l_df[1, "comp"], colnames(dms_add), sep = "_")
dms_add$id <- rownames(dms_add)

m5C_prop_dms_df <- left_join(m5C_prop_dms_df, dms_add, by = "id")

table(m5C_prop_dms_df$SCC_class)
rm(dms_add)
```

```{r}
dms_prop_ids <- m5C_prop_dms_df |> filter(SCC_class != "=") |> pull(id) |> unique()
#top20_dms_ids_pval <- m5C_prop_dms_l_df |> arrange(p_val) |> slice_head(n = 20) |> pull(id)
```

```{r plot of DMS m5Cs - counts}
in_bp <- m5C_prop_dms_df |> filter(SCC_class != "=")

pDEGs<-plot_DEGs(in_bp,
          contrast_col="SCC_comp", 
          color_vec=c("#5e548e","#6b9080"),
          class_col="SCC_class")

#"#F21A00","#f7b32b", "#046C9A", "#273046"

pDEGs
ggsave(plot = pDEGs, file = "./Dorado_V1/m5C/peak_seq_C/m5C_DMS_prop.pdf", device = cairo_pdf, width = 5, height = 4)
```

#### m5C modification counts

```{r DMS m5C modification counts with EdgeR}
design <- model.matrix(~0+condition, data = edge_c$samples) # create a binary matrix to pair replicates (description of the conditions)
colnames(design) <- unique(edge_c$samples$condition)
design

edge_d <- estimateDisp(edge_n, design, robust = T) # robust = TRUE to have a better estimation of the outliers

edge_f <- glmQLFit(edge_d, design, robust = T) #squeeze of the variances -> estimation of the quasi-likelihood (QL) dispersions around

```

```{r}
edgeRglmQLF_2022 <- function(glmfit, # object of class DGEGLM
                      contrast_chr, # comparison (string)
                      comp_name="", # include in label column
                      signal_col="CPM", # signal column
                      signal_thr=0.1, # signal threshold
                      fc_col="log2_FC",
                      fc_thr=0, # fold change threshold
                      pval_col="p_val", # p-value column
                      pval_thr=0.05) # p-value threshold
                      { 
  
  contro<-makeContrasts(contrasts=contrast_chr,levels=glmfit$design)
  
  degs<-glmQLFTest(glmfit,
                   contrast= contro)$table[,c("logFC","PValue")]
  colnames(degs)<-c("log2_FC","p_val")
  
  degs$id<-rownames(degs)
  degs$comp<-ifelse(comp_name!="",comp_name,contrast_chr)
  degs$p_adj<-p.adjust(degs$p_val, method ="BH")
  
  active_levels<-rownames(contro)[which(contro!=0)]
  active_samples<-which(glmfit$samples$group%in%active_levels)
  cpm_sele<-cpm(glmfit,log=F)[,active_samples]
  
  degs$CPM<-round(rowMeans(cpm_sele),3)
  #degs$CPM<-round(2^degs_u$log2_CPM,3)
  
  degs$class<-"="
  degs[which(degs[,signal_col]>=signal_thr & 
             degs[,fc_col]>=fc_thr & 
             degs[,pval_col]<=pval_thr),"class"]<-"+"

  degs[which(degs[,signal_col]>=signal_thr &
             degs[,fc_col]<=(-fc_thr) &
             degs[,pval_col]<=pval_thr),"class"]<-"-"
  
  degs<-degs %>% dplyr::select(id,comp,class,log2_FC,p_val,p_adj,CPM)

  return(degs)
}

```


```{r EdgeR DNS extraction of results}
# extract results
m5C_dms_counts_w_df <- NULL
m5C_dms_counts_l_df <- NULL

contro_list<-list("SCC"="SCC-HK")
  
m5C_dms_counts_l_df <- edgeRglmQLF_2022(glmfit = edge_f,
                             contrast_chr = contro_list$SCC,
                             comp_name = names(contro_list),
                             signal_col = "CPM", signal_thr = -Inf,
                             fc_col = "log2_FC", fc_thr = 0.5,
                             pval_col = "p_val", pval_thr = 0.05)
  
#rownames(m5C_dms_counts_l_df) <- NULL
  
dms_add <- m5C_dms_counts_l_df |> dplyr::select("class", "log2_FC", "p_val", "p_adj", "CPM")
colnames(dms_add) <- paste(m5C_dms_counts_l_df[1, "comp"], colnames(dms_add), sep = "_")
dms_add$id <- rownames(dms_add)
dms_add <- relocate(dms_add, id)

m5C_dms_counts_w_df <- dms_add
  
#table(m5C_dms_counts_w_df$SCC_class)
rm(dms_add)
```

```{r plot of DMS m5Cs - counts}
in_bp <- m5C_dms_counts_l_df |> filter(class != "=")

pDEGs<-plot_DEGs(in_bp,
          contrast_col="comp", 
          color_vec=c("#5e548e","#6b9080"),
          class_col="class")

#"#F21A00","#f7b32b", "#046C9A", "#273046"

pDEGs
ggsave(plot = pDEGs, file = "./Dorado_V1/m5C/peak_seq_C/m5C_DMS_mod_counts.pdf", device = cairo_pdf, width = 5, height = 4)
```

```{r}
## adding info on counts DMS
m5C_r_anno <- left_join(m5C_r_anno, m5C_dms_counts_w_df |> 
                          dplyr::select(id,SCC_class, SCC_log2_FC, SCC_p_val), by = "id") |> 
                          dplyr::rename("counts_SCC_class" = "SCC_class", "counts_SCC_log2_FC" = "SCC_log2_FC", "counts_SCC_p_val" = "SCC_p_val") 

## adding info on proportions DMS
m5C_r_anno <- left_join(m5C_r_anno, m5C_prop_dms_df |> 
                          dplyr::select(id,SCC_class, SCC_logFC, SCC_p_val), by = "id") |> 
                          dplyr::rename("prop_SCC_class" = "SCC_class", "prop_SCC_log2_FC" = "SCC_logFC", "prop_SCC_p_val" = "SCC_p_val")

m5C_r_anno <- left_join(m5C_r_anno, r_anno_complete |>
                          dplyr::select(id,new_region_class))
```

```{r}
### computing combined p-value between proportions and counts
library(metap)
combined_p_value <- m5C_r_anno |> dplyr::select(id, prop_SCC_p_val,counts_SCC_p_val) |> 
  pivot_longer(2:3, names_to = "measure_class", values_to = "pval")|> 
  dplyr::mutate(combined_pval = sumlog(pval)$p, .by = (id))

m5C_r_anno <- left_join(m5C_r_anno, combined_p_value |> dplyr::select(id, combined_pval), by = "id")
m5C_r_anno <- m5C_r_anno |> unique()

m5C_r_anno_clip <- m5C_r_anno |> dplyr::filter(!is.na(eclip_cor_class))
```

### Intersection DMS

```{r}
### detection of sites with inverse behavior between proportions and counts differential analysis
m5C_r_anno$prop_SCC_nopval_class <- "="
m5C_r_anno$prop_SCC_nopval_class[which(m5C_r_anno$prop_SCC_log2_FC > 0)] <- "+"
m5C_r_anno$prop_SCC_nopval_class[which(m5C_r_anno$prop_SCC_log2_FC < 0)] <- "-"

table(m5C_r_anno$prop_SCC_nopval_class)

m5C_r_anno$inverse <- ifelse((m5C_r_anno$prop_SCC_nopval_class == "+" & m5C_r_anno$counts_SCC_class != "+"|m5C_r_anno$prop_SCC_nopval_class == "-" & m5C_r_anno$counts_SCC_class != "-"), T, F)
table(m5C_r_anno$inverse)

m5C_r_anno$combined_class <- "Equally_methylated"
m5C_r_anno$combined_class[which(m5C_r_anno$counts_SCC_log2_FC >= 0.5 & m5C_r_anno$combined_pval <= 0.05 & !m5C_r_anno$inverse == T)] <- "SCC_up"
m5C_r_anno$combined_class[which(m5C_r_anno$counts_SCC_log2_FC <= -0.5 & m5C_r_anno$combined_pval <= 0.05 & !m5C_r_anno$inverse == T)] <- "HK_up"

m5C_r_anno_clip <- m5C_r_anno |> dplyr::filter(!is.na(eclip_cor_class))

table(m5C_r_anno$combined_class)
```

```{r}
scatter_df <- m5C_r_anno #|> unique()
class(scatter_df$combined_pval)

ggplot(scatter_df, aes(x = counts_SCC_log2_FC, y = -log10(combined_pval), fill = combined_class, color = combined_class)) + 
  geom_point(alpha=0.8) + 
  theme_bw() + 
  scale_color_manual(values = c("grey50","#6b9080","#5e548e")) + 
  geom_hline(yintercept = -log10(0.05), colour = "grey",linetype = "dashed") + 
  geom_vline(xintercept = 0.5,colour = "grey",linetype = 2) +
  geom_vline(xintercept = -0.5,colour = "grey", linetype = 2)

#ggsave("./Dorado_V1/m5C/volcano_dms_combined_class.png", device = "png", width = 5, height = 4)
```

```{r}
scatter_df$combined_class <- factor(scatter_df$combined_class, levels = c("Equally_methylated", "HK_up", "SCC_up"))
scatter_df <- arrange(scatter_df, combined_class)

ggplot(scatter_df, aes(x = prop_SCC_log2_FC, y = counts_SCC_log2_FC, color=combined_class)) +
  geom_point(alpha=0.8) + 
  theme_bw() + 
  scale_color_manual(values = c("#a6a6a6","#40ab7c","#ae68ff")) + 
  geom_hline(yintercept = 0.5, colour = "grey",linetype = "dashed") +
  geom_hline(yintercept = -0.5, colour = "grey",linetype = "dashed") +
  geom_vline(xintercept = 0,colour = "grey",linetype = 2) +
  theme(legend.position = "none")

#ggsave("./Dorado_V1/m5C/peak_seq_C/scatter_counts_prop_log2FC.png", device = "png", width = 4, height = 4)
```

#### DMS transcript localization 

```{r}
in_bp <- m5C_r_anno |> 
  filter(combined_class != "Equally_methylated") |> 
  summarise(N_m5C_per_region = length(unique(id)), .by = c("combined_class", "new_region_class")) |>
  mutate(tot = sum(N_m5C_per_region), .by = "combined_class") |>
  mutate(perc = (N_m5C_per_region/tot)*100, .by = "combined_class")
  
in_bp$new_region_class <- factor(in_bp$new_region_class, levels = c("3_UTR", "CDS", "5_UTR", "Exon", "Intron", "Intergenic"))

ggplot(in_bp, aes(x = new_region_class, y = perc, fill = combined_class)) + 
  geom_bar(stat = "identity") +
  facet_wrap(~combined_class) +
  theme_tufte(base_size = bs, base_family = bf) +
  coord_flip() +
  scale_fill_manual(values = rev(c("#5e548e","#6b9080"))) +
  theme(legend.position = "none", axis.title.y = element_blank()) +
  labs(y = "% m5Cs") +
  #labs(y = "Number m5Cs") +
  geom_text(aes(label = paste(round(perc,1), "%"), hjust = ifelse(in_bp$new_region_class %in% c("3_UTR", "CDS"),1,-0.1)))

#ggsave("./Dorado_V1/m5C/peak_seq_C/m5C_transcript_localization.pdf", device = cairo_pdf, width = 5, height = 4)
#write.xlsx(in_bp, file="~/Desktop/Nanopore_plots_data/2f_m5C_transcript_localization.xlsx")
```

```{r}
in_bp <- m5C_r_anno |> dplyr::filter(combined_class != "Equally_methylated") |>
  summarise(N_m5C_per_gene = length(unique(id)), .by = c("external_gene_name", "combined_class")) |> 
  mutate(N_genes = length(unique(external_gene_name)), .by = c("combined_class", "N_m5C_per_gene"))

in_bp$N_m5C_plot <- in_bp$N_m5C_per_gene
in_bp$N_m5C_plot[which(in_bp$N_m5C_per_gene >= 1 & in_bp$N_m5C_per_gene <= 5)] <- "1-5"
in_bp$N_m5C_plot[which(in_bp$N_m5C_per_gene > 5 & in_bp$N_m5C_per_gene <= 10)] <- "5-10"
in_bp$N_m5C_plot[which(in_bp$N_m5C_per_gene > 10 & in_bp$N_m5C_per_gene <= 15)] <- "10-15"
in_bp$N_m5C_plot[which(in_bp$N_m5C_per_gene > 15 & in_bp$N_m5C_per_gene <= 20)] <- "15-20"
in_bp$N_m5C_plot[which(in_bp$N_m5C_per_gene > 20 & in_bp$N_m5C_per_gene <= 25)] <- "20-25"
in_bp$N_m5C_plot[which(in_bp$N_m5C_per_gene > 25 & in_bp$N_m5C_per_gene <= 30)] <- "25-30"
in_bp$N_m5C_plot[which(in_bp$N_m5C_per_gene > 30 & in_bp$N_m5C_per_gene <= 35)] <- "30-35"
in_bp$N_m5C_plot[which(in_bp$N_m5C_per_gene > 35 & in_bp$N_m5C_per_gene <= 40)] <- "35-40"
in_bp$N_m5C_plot[which(in_bp$N_m5C_per_gene > 40 & in_bp$N_m5C_per_gene <= 45)] <- "40-45"
in_bp$N_m5C_plot[which(in_bp$N_m5C_per_gene > 45 & in_bp$N_m5C_per_gene <= 50)] <- "45-50"

in_bp$N_m5C_plot[which(in_bp$N_m5C_per_gene > 50)] <- ">50"
in_bp$N_genes_plot <- in_bp$N_genes

in_bp <- in_bp |> na.omit() |> group_by(combined_class,N_m5C_plot) |> mutate(N_genes_try = length(unique(external_gene_name)))

in_bp$N_m5C_plot <- factor(in_bp$N_m5C_plot, levels = c("1-5", "5-10", "10-15", "15-20", "20-25",
                                                        "25-30", "30-35", "35-40", "40-45", "45-50" ,">50"))

in_bp <- in_bp |> dplyr::select(N_m5C_plot, N_genes_try, combined_class) |> unique()

ggplot(in_bp, aes(x = N_m5C_plot, y = N_genes_try, fill = combined_class)) + 
  geom_bar(stat = "identity", position = "dodge") +
  geom_text(aes(label = N_genes_try), angle = 90, hjust = ifelse(in_bp$N_m5C_plot == "1-5",1,-0.2), color = "black", position = position_dodge(0.9)) +
  #scale_y_cut(breaks = 1100, expand = F, scales = 0.5) +
  scale_fill_manual(values = rev(c("#5e548e","#6b9080"))) +
  labs(x = "Number m5Cs", y = "Number of genes") +
  theme_bw() +
  theme(legend.position = c("top"),
        legend.title = element_blank(),
        panel.grid.minor.y = element_blank()) #+
  labs(title = "Number of m5Cs per transcript")
  expand_limits(y = 4000)

#ggsave("./Dorado_V1/m5C/peak_seq_C/genes_classification_per_m5C_general_50_num.pdf", device = cairo_pdf, width = 5, height = 4)
```

### Binned barplot of proportion 

```{r}
m5C_prop_l_df <- m5C_prop |> mutate(id = rownames(m5C_prop)) |> pivot_longer(-id, names_to = "sample", values_to = "proportion")
m5C_prop_l_df <- left_join(m5C_prop_l_df, c_anno)

m5C_mean_prop <- m5C_prop_l_df |> group_by(id,condition) |> dplyr::summarise(mean_prop = mean(proportion)) |> dplyr::mutate(log_mean_prop = log2(mean_prop)) |> unique()

m5C_mean_prop$bin <- "0"
m5C_mean_prop$bin[which(m5C_mean_prop$mean_prop >= 0 & m5C_mean_prop$mean_prop <= 0.1)] <- "0-0.1" 
m5C_mean_prop$bin[which(m5C_mean_prop$mean_prop > 0.1 & m5C_mean_prop$mean_prop <= 0.2)] <- "0.1-0.2"
m5C_mean_prop$bin[which(m5C_mean_prop$mean_prop > 0.2 & m5C_mean_prop$mean_prop <= 0.3)] <- "0.2-0.3"
m5C_mean_prop$bin[which(m5C_mean_prop$mean_prop > 0.3 & m5C_mean_prop$mean_prop <= 0.4)] <- "0.3-0.4"
m5C_mean_prop$bin[which(m5C_mean_prop$mean_prop > 0.4 & m5C_mean_prop$mean_prop <= 0.5)] <- "0.4-0.5"
m5C_mean_prop$bin[which(m5C_mean_prop$mean_prop > 0.5 & m5C_mean_prop$mean_prop <= 0.6)] <- "0.5-0.6"
m5C_mean_prop$bin[which(m5C_mean_prop$mean_prop > 0.6 & m5C_mean_prop$mean_prop <= 0.7)] <- "0.6-0.7"
m5C_mean_prop$bin[which(m5C_mean_prop$mean_prop > 0.7 & m5C_mean_prop$mean_prop <= 0.8)] <- "0.7-0.8"
m5C_mean_prop$bin[which(m5C_mean_prop$mean_prop > 0.8 & m5C_mean_prop$mean_prop <= 0.9)] <- "0.8-0.9"
m5C_mean_prop$bin[which(m5C_mean_prop$mean_prop > 0.9 & m5C_mean_prop$mean_prop <= 1)] <- "0.9-1"

m5C_mean_prop <- left_join(m5C_mean_prop,  m5C_r_anno |> dplyr::select(id, combined_class))
```

```{r}
m5C_mean_prop$bin <- factor(m5C_mean_prop$bin, levels = c("0-0.1", "0.1-0.2", "0.2-0.3", "0.3-0.4", "0.4-0.5","0.5-0.6", 
                                          "0.6-0.7", "0.7-0.8", "0.8-0.9", "0.9-1"))
m5C_mean_prop$condition <- paste0(m5C_mean_prop$condition, "_up")
in_bp <- m5C_mean_prop |> dplyr::filter(condition == combined_class) |> 
  dplyr::select(-condition) |> unique() |> ungroup() |>
  dplyr::filter(combined_class != "Equally_methylated") |> group_by(combined_class, bin) |> summarise(N = length(unique(id)))

#in_bp <- m5C_mean_prop |> dplyr::filter(condition == dms_general_class & id %in% m5C_r_anno_clip$id) #|> dplyr::select(-condition) |> unique() |> ungroup() |> filter(dms_general_class != "Equal") |> group_by(dms_general_class, bin) |> summarise(N = length(unique(id)))

ggplot(in_bp, aes(x = bin, y = N, fill= combined_class)) +
  geom_bar(stat = "identity", position = "dodge") + 
  geom_text(aes(label = N), angle = 90, hjust = 1.2, color = "black", position = position_dodge(0.9)) +
  scale_fill_manual(values = rev(c("#5e548e","#6b9080"))) +
  theme_bw() + 
  scale_y_log10(breaks = c(10,100,1000,10000,100000)) +
  #scale_y_cut(breaks = c(11000), expand = F, space = 0.3) +
  #geom_text(aes(label = N), angle = 90) +
  theme(legend.position = c("top"),
        legend.title = element_blank(),
        panel.grid.minor.y = element_blank(),
        axis.text.x = element_text(angle = 45, vjust = 0.5)) +
  labs(y = "log10(N)")
  expand_limits(y = c(0,4500000))
  
#ggsave("./Dorado_V1/m5C/peak_seq_C/binned_stoichiometry.pdf", device = cairo_pdf, width = 5, height = 4)
#write.xlsx(in_bp, file = "~/Desktop/Nanopore_plots_data/2d_binned_stoichiometry.xlsx")
```

```{r}
save(filt_mod_df, prop_data, perc_df, m5C_dms_counts_l_df, m5C_dms_counts_w_df, m5C_r_anno, m5C_prop, m5C_prop_dms_df, m5C_prop_dms_l_df, r_anno, r_anno_complete, structure_df, c_anno, m5C_mod_norm_df, file = "./Dorado_V1/m5C/peak_seq_C/m5C_DMS.RData")
```

### Motif analysis

```{r}
#load("~/Desktop/anno/enst_seq.RData")

motif_m5C_r_anno <- m5C_r_anno
motif_m5C_r_anno$motif_start <- motif_m5C_r_anno$start - 2
motif_m5C_r_anno$motif_end <- motif_m5C_r_anno$end + 2

motif_m5C_granges <- makeGRangesFromDataFrame(motif_m5C_r_anno |> dplyr::select(-c(start,end)),
                                         keep.extra.columns = T,
                                         ignore.strand = F,
                                         seqinfo = NULL,
                                         seqnames.field = "chr",
                                         start.field = "motif_start",
                                         end.field = "motif_end",
                                         strand.field = "strand",
                                         starts.in.df.are.0based = T)

m5C_r_anno$motif_seq <- getSeq(BSgenome.Hsapiens.UCSC.hg38, motif_m5C_granges) |> as.character() |> stringr::str_replace_all("T", "U")
```

#### General logos

```{r}
library(ggseqlogo)

equally_met_motif_df <- m5C_r_anno |> 
  dplyr::filter(combined_class == "Equally_methylated" & peak_seq == "C") |> 
  dplyr::mutate(motif_seq = str_replace_all(motif_seq, "_", "-"))

ggseqlogo(equally_met_motif_df$motif_seq, method = "probability") + labs(title = "Equally Methylated m5C", subtitle = paste("( N°", length(equally_met_motif_df$motif_seq), ")", sep = " "))

#ggsave("./Dorado_V1/m5C/motif_analysis/Eq_met.pdf", device = cairo_pdf, width = 5, height = 4)
```

```{r}
SCC_motif_df <- m5C_r_anno |> 
  dplyr::filter(combined_class == "SCC_up" & peak_seq == "C") |> 
  dplyr::mutate(motif_seq = str_replace_all(motif_seq, "_", "-"))

ggseqlogo(SCC_motif_df$motif_seq, method = "probability") + labs(title = "SCC m5C", subtitle = paste("( N°", length(SCC_motif_df$motif_seq), ")", sep = " "))

#ggsave("./Dorado_V1/m5C/motif_analysis/SCC_met.pdf", device = cairo_pdf, width = 5, height = 4)
```

```{r}
HK_motif_df <- m5C_r_anno |> 
  dplyr::filter(combined_class == "HK_up"& peak_seq == "C") |> 
  dplyr::mutate(motif_seq = str_replace_all(motif_seq, "_", "-"))

ggseqlogo(HK_motif_df$motif_seq, method = "probability") + labs(title = "HK m5C", subtitle = paste("( N°", length(HK_motif_df$motif_seq), ")", sep = " "))

#ggsave("./Dorado_V1/m5C/motif_analysis/HK_met.pdf", device = cairo_pdf, width = 5, height = 4)
```

#### Logos with m5C within YBX1 regions

```{r}
YBX1_equally_met_motif_df <- m5C_r_anno |> 
  dplyr::filter(combined_class == "Equally_methylated" & !is.na(eclip_class) & peak_seq == "C") |> 
  dplyr::mutate(motif_seq = str_replace_all(motif_seq, "_", "-"))

ggseqlogo(YBX1_equally_met_motif_df$motif_seq, method = "probability") + labs(title = "Equally Methylated m5C in YBX1 binding regions", subtitle = paste("( N°", length(YBX1_equally_met_motif_df$motif_seq), ")", sep = " "))

#ggsave("./Dorado_V1/m5C/motif_analysis/Eq_met_YBX1.pdf", device = cairo_pdf, width = 5, height = 4)
```

```{r}
YBX1_SCC_motif_df <- m5C_r_anno |> 
  dplyr::filter(combined_class == "SCC_up" & !is.na(eclip_class) & peak_seq == "C") |> 
  dplyr::mutate(motif_seq = str_replace_all(motif_seq, "_", "-"))

ggseqlogo(YBX1_SCC_motif_df$motif_seq, method = "probability") + labs(title = "SCC m5C in YBX1 binding regions", subtitle = paste("( N°", length(YBX1_SCC_motif_df$motif_seq), ")", sep = " "))

#ggsave("./Dorado_V1/m5C/motif_analysis/SCC_met_YBX1.pdf", device = cairo_pdf, width = 5, height = 4)
```

```{r}
YBX1_HK_motif_df <- m5C_r_anno |> 
  dplyr::filter(combined_class == "HK_up" & !is.na(eclip_class)& peak_seq == "C") |> 
  dplyr::mutate(motif_seq = str_replace_all(motif_seq, "_", "-"))

ggseqlogo(YBX1_HK_motif_df$motif_seq, method = "probability") + labs(title = "HK m5C in YBX1 binding regions", subtitle = paste("( N°", length(YBX1_HK_motif_df$motif_seq), ")", sep = " "))

#ggsave("./Dorado_V1/m5C/motif_analysis/HK_met_YBX1.pdf", device = cairo_pdf, width = 5, height = 4)
```

#### Logos specific for m5C in same direction of YBX1 binding

```{r}
YBX1_SCC_motif_df <- m5C_r_anno |> 
  dplyr::filter(combined_class == "SCC_up" & eclip_class == "SCC_gain" & peak_seq == "C") |> 
  dplyr::mutate(motif_seq = str_replace_all(motif_seq, "_", "-"))

ggseqlogo(YBX1_SCC_motif_df$motif_seq, method = "probability") + labs(title = "SCC m5C in YBX1 SCC_gain regions", subtitle = paste("( N°", length(YBX1_SCC_motif_df$motif_seq), ")", sep = " "))

#ggsave("./Dorado_V1/m5C/motif_analysis/SCC_YBX1_SCC_gain.pdf", device = cairo_pdf, width = 5, height = 4)
```


```{r}
YBX1_HK_motif_df <- m5C_r_anno |> 
  dplyr::filter(combined_class == "HK_up" & eclip_class == "HK_gain" & peak_seq == "C") |> 
  dplyr::mutate(motif_seq = str_replace_all(motif_seq, "_", "-"))

ggseqlogo(YBX1_HK_motif_df$motif_seq, method = "probability") + labs(title = "HK m5C in YBX1 HK_gain regions", subtitle = paste("( N°", length(YBX1_HK_motif_df$motif_seq), ")", sep = " "))

#ggsave("./Dorado_V1/m5C/motif_analysis/HK_met_YBX1_HK_gain.pdf", device = cairo_pdf, width = 5, height = 4)
```

### Metaprofile

```{r}
DMS_m5C_ids <- m5C_r_anno |> filter(combined_class != "Equally_methylated") |> pull(id) |> unique()
DMS_m5C_YBX1_ids <- m5C_r_anno |> dplyr::filter(combined_class != "Equally methylated" & !is.na(eclip_class)) |>pull(id) |> unique() 

load("~/Desktop/Nanopore/Results/Dorado_V1/m5C/metaprofile_dat/HK_gain_m5C_coord_tx_df.RData")
load("~/Desktop/Nanopore/Results/Dorado_V1/m5C/metaprofile_dat/SCC_gain_m5C_coord_tx_df.RData")

m5C_DMS_coord_tx_df <- rbind(HK_gain_m5C_coord_tx_df, SCC_gain_m5C_coord_tx_df)

m5C_DMS_coord_tx_df <- m5C_DMS_coord_tx_df |>
  dplyr::rename("ensembl_transcript_id" = "tx_id", "ensembl_exon_id" = "exon_id") |>
  #dplyr::filter(peak_seq == "C") |>
  dplyr::select(id,start,ensembl_transcript_id,ensembl_exon_id,exon_rank,
                seq_strand,combined_class,region_class,peak_seq) |>
  dplyr::filter(id %in% DMS_m5C_ids & region_class != "Exon") |> ## filter to retain only sites with accordance in direction between proportions and counts
  dplyr::rename("zero-based_position" = "start", "strand" = "seq_strand") |> 
  dplyr::mutate("one-based_position" = `zero-based_position`+1) |>
  dplyr::relocate(`one-based_position`, .after = `zero-based_position`)

m5C_DMS_coord_tx_df <- left_join(m5C_DMS_coord_tx_df, m5C_r_anno |> dplyr::select(id, combined_pval, counts_SCC_log2_FC,clip_overlap,id_name, eclip_class))

m5C_DMS_YBX1_coord_tx_df <- m5C_DMS_coord_tx_df |> dplyr::filter(id %in% DMS_m5C_YBX1_ids)

m5C_YBX1_concordant_tx_df <- m5C_DMS_coord_tx_df |>
  dplyr::filter(eclip_class == "SCC_gain" & combined_class == "SCC_up" | eclip_class == "HK_gain" & combined_class == "HK_up")
```

```{r}
load("~/Desktop/anno/ens_anno_110_hs.RData")

## checking the general proportion of each transcript region
enst_slim <- enst |> 
  dplyr::filter(canonical == "T" & !l_5utr == 0 & !l_cds == 0 & !l_3utr == 0 & transcript_biotype == "protein_coding") |>
  mutate(perc_5utr = (l_5utr/transcript_length)*100,
         perc_cds = (l_cds/transcript_length)*100,
         perc_3utr = (l_3utr/transcript_length)*100, .by = "ensembl_transcript_id") |>
  #mutate(mean_perc_5utr = round(mean(perc_5utr)),
  #       mean_perc_cds = round(mean(perc_cds)),
  #       mean_perc_3utr =  round(mean(perc_3utr))) |>
  mutate(tx_5utr_start = 1,
         tx_5utr_end = (tx_5utr_start + (l_5utr))) |>
  mutate(tx_cds_start = tx_5utr_end+1,
         tx_cds_end = (tx_5utr_end + (l_cds))) |>
  mutate(tx_3utr_start = tx_cds_end + 1,
         tx_3utr_end = (tx_cds_end + (l_3utr)))

ense_slim <- ense |> dplyr::filter(ensembl_transcript_id %in% enst_slim$ensembl_transcript_id)

anno_df_slim <- enst_slim |> dplyr::select(ensembl_gene_id,ensembl_transcript_id,l_5utr,l_cds,l_3utr,tx_5utr_start,tx_5utr_end,tx_cds_start,tx_cds_end,tx_3utr_start,tx_3utr_end)
anno_df_slim <- left_join(anno_df_slim, ense |> dplyr::select(ensembl_transcript_id, ensembl_exon_id)) |> dplyr::relocate(ensembl_exon_id, .after = ensembl_transcript_id)
anno_df_slim <- left_join(anno_df_slim, ensg |> dplyr::select(ensembl_gene_id,external_gene_name))
```

```{r}
df <- m5C_YBX1_concordant_tx_df
anno_df <- anno_df_slim

split_by <- "combined_class"
total_bins = 50

percentages = T

create_m5C_metaprofile <- function(df,anno_df, total_bins = 100, target = "", split_by = "", percentages = F){
  
  library(tidyverse)
  library(ggplot2)
  
  split_by <- rlang::sym(split_by)
  
  ## default proportions
  prop_5utr <- 0.07
  prop_cds <- 0.52
  prop_3utr <- 0.41
  
  ## compute the number of bins for each transcript region
  bins_5utr <- round(total_bins * prop_5utr)
  bins_cds <- round(total_bins * prop_cds)
  bins_3utr <- total_bins - (bins_5utr + bins_cds)
  
  message(sprintf("Binning: 5'UTR = %d, CDS = %d, 3'UTR = %d", bins_5utr, bins_cds, bins_3utr))
  
  ## check required columns
  required_cols <- c("ensembl_transcript_id", "zero-based_position")
  required_anno_cols <- c("ensembl_transcript_id", "tx_5utr_start", "tx_5utr_end",
                          "tx_cds_start", "tx_cds_end", "tx_3utr_start", "tx_3utr_end", "l_3utr", "l_cds", "l_5utr")
  
  stopifnot(all(required_cols %in% colnames(df)))
  stopifnot(all(required_anno_cols %in% colnames(anno_df)))
  
  df <- inner_join(df, anno_df)
  
  df <- df |>
  dplyr::filter(!is.na(region_class)) |>
  dplyr::mutate(region = case_when(
    `zero-based_position` >= tx_5utr_start & `zero-based_position` <= tx_5utr_end ~ "5_UTR",
    `zero-based_position` >= tx_cds_start & `zero-based_position` <= tx_cds_end ~ "CDS",
    `zero-based_position` >= tx_3utr_start & `zero-based_position` <= tx_3utr_end ~ "3_UTR"
  )) |>
  dplyr::mutate(rel_pos = case_when(
    region == "5_UTR" ~ `zero-based_position` - tx_5utr_start+1,
    region == "CDS" ~ `zero-based_position` - tx_cds_start+1,
    region == "3_UTR" ~ `zero-based_position` - tx_3utr_start+1
  )) |>
  mutate(bin = case_when(
    region == "5_UTR" ~ ceiling((rel_pos / l_5utr) * bins_5utr),
    region == "CDS" ~ ceiling((rel_pos / l_cds) * bins_cds) + bins_5utr,
    region == "3_UTR" ~ ceiling((rel_pos / l_3utr) * bins_3utr) + bins_cds + bins_5utr
  )) |>
  mutate(bin = pmin(bin, total_bins))

    if(split_by != "" & target == ""){
      
      metaprofile <- df |> 
        group_by(bin, !!split_by) |> 
        summarise(m5C_count = length(unique(id)), .groups = "drop") |>
        complete(bin = 1:(total_bins),!!split_by ,fill = list(m5C_count = 0)) |>
        ungroup() |>
        mutate(perc = (m5C_count/(sum(m5C_count))), .by = !!split_by)
      
      annotation <- metaprofile |> 
        group_by(!!split_by) |> 
        summarise(N_tot = sum(m5C_count)) #|> 
        #mutate(x = max(metaprofile$bin - 1), y = max(metaprofile$m5C_count - 1))
      
      y_col <- rlang::sym(ifelse(percentages, "perc", "m5C_count"))
      
      gg_met <- ggplot(metaprofile, aes(x = bin, y = !!y_col, group = !!split_by, color = !!split_by)) +
      geom_point() +
      geom_density(stat = "identity") +
      geom_vline(xintercept = c(bins_5utr, bins_5utr + bins_cds), linetype = "dashed", color = "grey") + 
      scale_x_continuous(
        breaks = c(bins_5utr/2, (bins_5utr + bins_cds)/2, bins_cds+bins_3utr/2),
        labels = c("5'UTR", "CDS", "3'UTR")
      ) + 
      theme_bw() +
      theme(panel.grid.minor = element_blank(),
            panel.grid.major = element_blank(),
            legend.position = "top") + 
      scale_color_manual(values = rev(c("#5e548e","#6b9080"))) +
      annotate("text", label = paste("HK = ", annotation$N_tot[which(annotation$combined_class == "HK_up")]), 
               x = max(metaprofile$bin-(total_bins/10)), y = ifelse(y_col == "m5C_count",max(metaprofile$m5C_count), max(metaprofile$perc))) +
      annotate("text", label = paste("SCC = ", annotation$N_tot[which(annotation$combined_class == "SCC_up")]), 
               x = max(metaprofile$bin-(total_bins/10)), y = ifelse(y_col == "m5C_count",max(metaprofile$m5C_count - 8), max(metaprofile$perc-0.002))) +
      labs(color = "Condition")
    
      } else {
        
      metaprofile <- df |> 
        group_by(bin) |> 
        summarise(m5C_count = length(unique(id)), .groups = "drop") |>
        complete(bin = 1:(total_bins), fill = list(m5C_count = 0)) |>
        mutate(perc = (m5C_count/(sum(m5C_count))))

      annotation <- metaprofile |> ungroup() |> summarise(N_tot = sum(m5C_count)) |> mutate(x = max(metaprofile$bin - 1), y = max(metaprofile$m5C_count - 1))
    
      y_col <- rlang::sym(ifelse(percentages, "perc", "m5C_count"))
      
      gg_met <- ggplot(metaprofile, aes(x = bin, y = !!y_col)) +
        geom_point(color = "steelblue") +
        geom_density(stat = "identity", color = "steelblue") +
        geom_vline(xintercept = c(bins_5utr, bins_5utr + bins_cds), linetype = "dashed", color = "grey") + 
        scale_x_continuous(
          breaks = c(bins_5utr/2, (bins_5utr + bins_cds)/2, bins_cds+bins_3utr/2),
          labels = c("5'UTR", "CDS", "3'UTR")
        ) + 
        theme_bw() +
        theme(panel.grid.minor = element_blank(),
              panel.grid.major = element_blank(),
              legend.position = "top")
        
      }
  
#   in_bp <- df |> group_by(, region_class) |> summarise(N = length(unique(id_name))) |> ungroup() |> dplyr::mutate(perc = (N/sum(N))*100, .by = eclip_class)
# 
# in_bp$region_class <- factor(in_bp$region_class, levels = rev(c("5_UTR", "CDS","3_UTR")))
#   
#   ggloc <- ggplot(in_bp, aes(region_class,perc)) +
#   geom_bar(stat = "identity", position = position_dodge(preserve = "single"), fill = "#a8cca4") +
#   facet_wrap(~eclip_class)+
#   scale_x_discrete(drop = F)+
#   theme_tufte(base_size = bs, base_family = bf) +
#   theme(strip.text.x = element_text(size = 20)) +
#   theme(axis.title.y = element_blank()) +
#   labs(y = "Peaks (%)")+
#   geom_text(aes(label = round(perc,1)), hjust = ifelse(in_bp$region_class %in% c("CDS", "3_UTR"),1, -0.2)) +
#   #scale_y_continuous(labels = scales::label_number_si()) +
#   coord_flip()
  

  
  return(list(input_anno_df = df,metaprofile = metaprofile, metaprofile_plot = gg_met,
              bins = list(utr5 = bins_5utr, cds = bins_cds, utr3 = bins_3utr)))

}
```


```{r}
m5C_metaprofile <- create_m5C_metaprofile(m5C_DMS_coord_tx_df , anno_df = anno_df_slim, total_bins = 50, split_by = "combined_class", percentages = T)
m5C_metaprofile$metaprofile_plot

m5C_YBX1_metaprofile <- create_m5C_metaprofile(m5C_DMS_YBX1_coord_tx_df , anno_df = anno_df_slim, total_bins = 50, split_by = "combined_class", percentages = T)
m5C_YBX1_metaprofile$metaprofile_plot

m5C_YBX1_concordant_metaprofile <- create_m5C_metaprofile(m5C_YBX1_concordant_tx_df , anno_df = anno_df_slim, total_bins = 30, split_by = "combined_class", percentages = T)
m5C_YBX1_concordant_metaprofile$metaprofile_plot

#write.xlsx(m5C_metaprofile$metaprofile, file = "~/Desktop/2j_m5C_YBX1_metaprofile_50bins.xlsx")

#ggsave("./Dorado_V1/m5C/metaprofile_m5C_50bins.pdf", device = cairo_pdf, width = 5, height = 4)
#save(m5C_metaprofile, file =  "./Dorado_V1/m5C/m5C_metaprofile.RData")
```

### Overlap with eCLIP-seq YBX1-bound regions

#### DMS transcript localization 

```{r}
in_bp <- m5C_r_anno |> 
  filter(combined_class != "Equally_methylated" & !is.na(eclip_class)) |> 
  summarise(N_m5C_per_region = length(unique(id)), .by = c("combined_class", "new_region_class")) |>
  mutate(tot = sum(N_m5C_per_region), .by = "combined_class") |>
  mutate(perc = (N_m5C_per_region/tot)*100, .by = "combined_class")
  
in_bp$new_region_class <- factor(in_bp$new_region_class, levels = c("3_UTR", "CDS", "5_UTR", "Exon", "Intron", "Intergenic"))

ggplot(in_bp, aes(x = new_region_class, y = perc, fill = combined_class)) + 
  geom_bar(stat = "identity") +
  facet_wrap(~combined_class) +
  theme_tufte(base_size = bs, base_family = bf) +
  coord_flip() +
  scale_fill_manual(values = rev(c("#5e548e","#6b9080"))) +
  theme(legend.position = "none", axis.title.y = element_blank()) +
  labs(y = "% m5Cs") +
  #labs(y = "Number m5Cs") +
  geom_text(aes(label = paste(round(perc,1), "%"), hjust = ifelse(in_bp$new_region_class %in% c("3_UTR", "CDS"),1,-0.1)))

#ggsave("./Dorado_V1/m5C/YBX1_m5C_transcript_localization.pdf", device = cairo_pdf, width = 5, height = 4)
```

```{r}
in_bp <- m5C_r_anno |> dplyr::filter(combined_class != "Equally_methylated" & !is.na(eclip_class)) |>
  summarise(N_m5C_per_gene = length(unique(id)), .by = c("external_gene_name", "combined_class")) |> 
  mutate(N_genes = length(unique(external_gene_name)), .by = c("combined_class", "N_m5C_per_gene"))

in_bp$N_m5C_plot <- in_bp$N_m5C_per_gene
in_bp$N_m5C_plot[which(in_bp$N_m5C_per_gene >= 1 & in_bp$N_m5C_per_gene <= 5)] <- "1-5"
in_bp$N_m5C_plot[which(in_bp$N_m5C_per_gene > 5 & in_bp$N_m5C_per_gene <= 10)] <- "5-10"
in_bp$N_m5C_plot[which(in_bp$N_m5C_per_gene > 10 & in_bp$N_m5C_per_gene <= 15)] <- "10-15"
in_bp$N_m5C_plot[which(in_bp$N_m5C_per_gene > 15 & in_bp$N_m5C_per_gene <= 20)] <- "15-20"
in_bp$N_m5C_plot[which(in_bp$N_m5C_per_gene > 20 & in_bp$N_m5C_per_gene <= 25)] <- "20-25"
in_bp$N_m5C_plot[which(in_bp$N_m5C_per_gene > 25 & in_bp$N_m5C_per_gene <= 30)] <- "25-30"
in_bp$N_m5C_plot[which(in_bp$N_m5C_per_gene > 30 & in_bp$N_m5C_per_gene <= 35)] <- "30-35"
in_bp$N_m5C_plot[which(in_bp$N_m5C_per_gene > 35 & in_bp$N_m5C_per_gene <= 40)] <- "35-40"
in_bp$N_m5C_plot[which(in_bp$N_m5C_per_gene > 40 & in_bp$N_m5C_per_gene <= 45)] <- "40-45"
in_bp$N_m5C_plot[which(in_bp$N_m5C_per_gene > 45 & in_bp$N_m5C_per_gene <= 50)] <- "45-50"

in_bp$N_m5C_plot[which(in_bp$N_m5C_per_gene > 50)] <- ">50"
in_bp$N_genes_plot <- in_bp$N_genes

in_bp <- in_bp |> na.omit() |> group_by(combined_class,N_m5C_plot) |> mutate(N_genes_try = length(unique(external_gene_name)))
                                              "15-18", "18-21", "21-24", "24-27", "27-30" ,">30"))

in_bp$N_m5C_plot <- factor(in_bp$N_m5C_plot, levels = c("1-5", "5-10", "10-15", "15-20", "20-25",
                                                        "25-30", "30-35", "35-40", "40-45", "45-50" ,">50"))

in_bp <- in_bp |> dplyr::select(N_m5C_plot, N_genes_try, combined_class) |> unique()

ggplot(in_bp, aes(x = N_m5C_plot, y = N_genes_try, fill = combined_class)) + 
  geom_bar(stat = "identity", position = "dodge") +
  #geom_text(aes(label = N_genes_try), angle = 90, hjust = 1, color = "black", position = position_dodge(0.9)) +
  scale_y_cut(breaks = c(800,15), expand = F, scales = 0.5) +
  scale_fill_manual(values = rev(c("#5e548e","#6b9080"))) +
  labs(x = "Number m5Cs", y = "Number of genes") +
  theme_bw() +
  theme(legend.position = c("top"),
        legend.title = element_blank(),
        panel.grid.minor.y = element_blank()) #+
  #expand_limits(y = 650)

#ggsave("./Dorado_V1/m5C/YBX1_genes_classification_per_m5C_general_50_nonum.pdf", device = cairo_pdf, width = 5, height = 4)
```

### Binned barplot of proportion 

```{r}
m5C_prop_l_df <- m5C_prop |> mutate(id = rownames(m5C_prop)) |> pivot_longer(-id, names_to = "sample", values_to = "proportion")
m5C_prop_l_df <- left_join(m5C_prop_l_df, c_anno)

m5C_mean_prop <- m5C_prop_l_df |> group_by(id,condition) |> dplyr::summarise(mean_prop = mean(proportion)) |> dplyr::mutate(log_mean_prop = log2(mean_prop)) |> unique()

m5C_mean_prop$bin <- "0"
m5C_mean_prop$bin[which(m5C_mean_prop$mean_prop >= 0 & m5C_mean_prop$mean_prop <= 0.1)] <- "0-0.1" 
m5C_mean_prop$bin[which(m5C_mean_prop$mean_prop > 0.1 & m5C_mean_prop$mean_prop <= 0.2)] <- "0.1-0.2"
m5C_mean_prop$bin[which(m5C_mean_prop$mean_prop > 0.2 & m5C_mean_prop$mean_prop <= 0.3)] <- "0.2-0.3"
m5C_mean_prop$bin[which(m5C_mean_prop$mean_prop > 0.3 & m5C_mean_prop$mean_prop <= 0.4)] <- "0.3-0.4"
m5C_mean_prop$bin[which(m5C_mean_prop$mean_prop > 0.4 & m5C_mean_prop$mean_prop <= 0.5)] <- "0.4-0.5"
m5C_mean_prop$bin[which(m5C_mean_prop$mean_prop > 0.5 & m5C_mean_prop$mean_prop <= 0.6)] <- "0.5-0.6"
m5C_mean_prop$bin[which(m5C_mean_prop$mean_prop > 0.6 & m5C_mean_prop$mean_prop <= 0.7)] <- "0.6-0.7"
m5C_mean_prop$bin[which(m5C_mean_prop$mean_prop > 0.7 & m5C_mean_prop$mean_prop <= 0.8)] <- "0.7-0.8"
m5C_mean_prop$bin[which(m5C_mean_prop$mean_prop > 0.8 & m5C_mean_prop$mean_prop <= 0.9)] <- "0.8-0.9"
m5C_mean_prop$bin[which(m5C_mean_prop$mean_prop > 0.9 & m5C_mean_prop$mean_prop <= 1)] <- "0.9-1"

m5C_mean_prop <- left_join(m5C_mean_prop,  m5C_r_anno |> dplyr::select(id, combined_class))
```


```{r}
m5C_mean_prop$bin <- factor(m5C_mean_prop$bin, levels = c("0-0.1", "0.1-0.2", "0.2-0.3", "0.3-0.4", "0.4-0.5","0.5-0.6", 
                                          "0.6-0.7", "0.7-0.8", "0.8-0.9", "0.9-1"))
m5C_mean_prop$condition <- paste0(m5C_mean_prop$condition, "_up")

in_bp <- m5C_mean_prop[which(m5C_mean_prop$id %in% m5C_r_anno_clip$id),]
in_bp <- in_bp |>  dplyr::filter(condition == combined_class) |>
  #dplyr::filter(id %in% m5C_r_anno_clip$id) |> 
  dplyr::select(-condition) |> unique() |> ungroup() |> group_by(combined_class, bin) |> summarise(N = length(unique(id)))

ggplot(in_bp, aes(x = bin, y = N, fill= combined_class)) +
  geom_bar(stat = "identity", position = "dodge") + 
  #geom_text(aes(label = N), angle = 90, hjust = 1.2, color = "black", position = position_dodge(0.9)) +
  scale_fill_manual(values = rev(c("#5e548e","#6b9080"))) +
  theme_bw() + 
  scale_y_log10(breaks = c(10,100,1000,10000,100000)) +
  #scale_y_cut(breaks = c(11000), expand = F, space = 0.3) +
  #geom_text(aes(label = N))
  theme(legend.position = c("top"),
        legend.title = element_blank(),
        panel.grid.minor.y = element_blank(),
        axis.text.x = element_text(angle = 45, vjust = 0.5)) +
  labs(y = "log10(N)")
  expand_limits(y = c(0,4500000))
  
ggsave("./Dorado_V1/binned_combined_DMS_log10_nonum.pdf", device = cairo_pdf, width = 5, height = 4)
```

### Structural analysis

```{r}
structure_df <- read.xlsx("./Dorado_V1/m5C/YBX1_Table_structural_calculations.xlsx", sheet = 2)
structure_df <- structure_df[,-1]

## adding the new column of the transcript localization annotation
structure_df <- left_join(structure_df, m5C_r_anno |> dplyr::select(id,new_region_class))

#write.xlsx(structure_df, file="~/Desktop/YBX1_Sup_Data/SupData4_m5C_structure_pred.xlsx")


regions <- c("CDS", "3_UTR")
targets <- c("NRP1", "NSD2", "FOXK1")

p_val <- structure_df |> dplyr::filter(region_class %in%regions) |> #group_by(new_region_class) |>
  wilcox_test(`%dsG_WT`~m5C_combined_class) |>
  adjust_pvalue() |> 
  add_significance("p") |> 
  add_xy_position(x = "m5C_combined_class") 

ggplot(structure_df |> dplyr::filter(new_region_class %in%regions), 
       aes(x = m5C_combined_class, y = `%dsG_WT`, fill = m5C_combined_class, color = m5C_combined_class)) + 
  geom_violin(alpha = 0.1) + 
  #geom_beeswarm() +
  #facet_wrap(~new_region_class) + 
  geom_boxplot(notch = T, width = 0.2, fill = "white", outliers = F) + 
  scale_fill_manual(values = c(colors[3], colors[2])) +
  scale_color_manual(values = c(colors[3], colors[2])) + 
  theme_bw(base_family = bf, base_size = 15) + 
  theme(panel.grid.major.x = element_blank(),
        legend.position = "none") + 
  #stat_pvalue_manual(p_val, label = "p", step.increase = 0.01, hide.ns = T, inherit.aes = F) +
  coord_cartesian(ylim = c(40,75))
  geom_text_repel(aes(label = ifelse(external_gene_name %in% targets, external_gene_name, "")),
                  color = "black", max.overlaps = 10000000,min.segment.length = 0, vjust = 0.5, hjust = 0.5)

#ggsave("./Dorado_V1/m5C/dsG_HK_SCC_new_region_class.pdf", device = cairo_pdf, width = 3, height = 4.5)
```

```{r}
p_val <- structure_df |> dplyr::filter(region_class %in%regions) |> #group_by(new_region_class) |>
  wilcox_test(En_WT~m5C_combined_class) |>
  adjust_pvalue() |> 
  add_significance("p") |> 
  add_xy_position(x = "m5C_combined_class") 

ggplot(structure_df |> dplyr::filter(new_region_class %in%regions), 
       aes(x = m5C_combined_class, y = En_WT , fill = m5C_combined_class, color = m5C_combined_class)) + 
  geom_violin(alpha = 0.1) + 
  #geom_beeswarm() +
  #facet_wrap(~new_region_class) + 
  geom_boxplot(notch = T, width = 0.2, fill = "white", outliers = F) + 
  scale_fill_manual(values = c(colors[3], colors[2])) +
  scale_color_manual(values = c(colors[3], colors[2])) + 
  theme_bw(base_size = 15) + 
  theme(panel.grid.major.x = element_blank(),
        legend.position = "none") #+ 
  #stat_pvalue_manual(p_val, label = "p", step.increase = 0.03, hide.ns = T, inherit.aes = F) #+
  geom_text_repel(aes(label = ifelse(external_gene_name %in% targets, external_gene_name, "")), 
                    color = "black", max.overlaps = 10000000, min.segment.length = 0, vjust = 0.5, hjust = 0.5)

#ggsave("./Dorado_V1/m5C/En_HK_SCC_new_region_class.pdf", device = cairo_pdf, width = 3, height = 4.5)

write.xlsx(structure_df |>
             dplyr::filter(new_region_class %in%regions) |> 
             dplyr::select(m5C_combined_class, En_WT, m5C_combined_class, external_gene_name),
           file = "~/Desktop/Nanopore_plots_data/2l_FoldingEnergy_violin.xlsx")
```

### Functional enrichment

```{r}
library(enrichR)
```

```{r enrichment_functions,  warning=F, message=F, echo=F,eval=T}

### The function enrichment_enrichr performs enrichment analysis with the enrichR bioconductor library, interface to the enrichR webserver ---- 

enrichment_enrichr <- function(input_vec, # vector with gene names
                               input_name="i1", # name for input list
                               dbs_vec="", # vector with database names
                               return_size_in_anno = T # use only genes with at least 1 annotation to calculate input size
){ 
   
   input_vec <- input_vec %>% stringr::str_trim() %>% stringr::str_sort() %>% 
      unique() # clean input vector
   input_vec <- input_vec[!duplicated(stringr::str_to_upper(input_vec))] # remove case duplicates
   
   #library(tidyverse)
   #library(enrichR)
   #library(plyr)
   
   dbs_transcription <- c("ChEA_2022",
                          "ARCHS4_TFs_Coexp",
                          "TF_Perturbations_Followed_by_Expression",
                          "lncHUB_lncRNA_Co-Expression",
                          "FANTOM6_lncRNA_KD_DEGs",
                          "TRANSFAC_and_JASPAR_PWMs",
                          "Epigenomics_Roadmap_HM_ChIP-seq",
                          "TargetScan_microRNA_2017",
                          "miRTarBase_2017",
                          "ENCODE_TF_ChIP-seq_2015",
                          "TF-LOF_Expression_from_GEO",
                          "ENCODE_Histone_Modifications_2015",
                          "TRRUST_Transcription_Factors_2019")
   
   dbs_pathway<-c( "KEGG_2021_Human",
                   "KEGG_2019_Mouse",
                   "WikiPathways_2021_Human",
                   "WikiPathways_2019_Mouse",
                   "BioPlanet_2019",
                   "Reactome_2022",
                   "BioCarta_2016",
                   "Humancyc_2016",
                   "NCI-Nature_2016",
                   "Panther_2016",
                   "BioPlex_2017",
                   "huMAP",
                   "LINCS_L1000_Kinase_Perturbations_up",
                   "LINCS_L1000_Kinase_Perturbations_down",
                   "ARCHS4_Kinases_Coexp",
                   "Virus-Host_PPI_P-HIPSTer_2020",
                   "NURSA_Human_Endogenous_Complexome",
                   "Elsevier_Pathway_Collection",
                   "MSigDB_Hallmark_2020", # MSigDB hallmark collection
                   "SubCell_BarCode", # web portal to proteome subcellular localization https://www.subcellbarcode.org/
                   "The_Kinase_Library_2023",
                   "Metabolomics_Workbench_Metabolites_2022",
                   "HMS_LINCS_KinomeScan"
                   )
   
   dbs_onto<-c("GO_Biological_Process_2025",
               "GO_Cellular_Component_2025",
               "GO_Molecular_Function_2025",
               "MGI_Mammalian_Phenotype_Level_4_2021",
               "Jensen_COMPARTMENTS",
               "SynGO_2022",
               "KOMP2_Mouse_Phenotypes_2022"
               )
   
   dbs_disease<-c( "OMIM_Disease",
                   "OMIM_Expanded",
                   "Human_Phenotype_Ontology",
                   "Jensen_DISEASES",
                   "Achilles_fitness_decrease",
                   "Achilles_fitness_increase",
                   "DSigDB",
                   "MSigDB_Oncogenic_Signatures",
                   "PheWeb_2019",
                   "ClinVar_2019",
                   "GWAS_Catalog_2023",
                   "DisGeNET",
                   "ARCHS4_IDG_Coexp",
                   "dbGaP",
                   "Rare_Diseases_GeneRIF_Gene_Lists",
                   "Rare_Diseases_AutoRIF_Gene_Lists",
                   "Rare_Diseases_AutoRIF_ARCHS4_Predictions",
                   "Rare_Diseases_GeneRIF_ARCHS4_Predictions",
                   "COVID-19_Related_Gene_Sets_2021",
                   "TG_GATES_2020",
                   "Orphanet_Augmented_2021",
                   "PhenGenI_Association_2021",
                   "GTEx_Aging_Signatures_2021",
                   "Proteomics_Drug_Atlas_2023",
                   "GeDiPNet_2023",
                   "MAGMA_Drugs_and_Diseases",
                   "IDG_Drug_Targets_2022"
                    )
   
   dbs_tissue<-c( "Jensen_TISSUES",
                  "Human_Gene_Atlas",
                  "Mouse_Gene_Atlas",
                  "Allen_Brain_Atlas_up",
                  "Allen_Brain_Atlas_down",
                  "GTEx_Tissue_Sample_Gene_Expression_Profiles_up",
                  "GTEx_Tissue_Sample_Gene_Expression_Profiles_down",
                  "Cancer_Cell_Line_Encyclopedia",
                  "NCI-60_Cancer_Cell_Lines",
                  "Tissue_Protein_Expression_from_Human_Proteome_Map",
                  "ARCHS4_Tissues",
                  "ESCAPE",
                  "CCLE_Proteomics_2020",
                  "ProteomicsDB_2020",
                  "SysMyo_Muscle_Gene_Sets",
                  "Descartes_Cell_Types_and_Tissue_2021",
                  "Allen_Brain_Atlas_10x_scRNA_2021",
                  "Azimuth_Cell_Types_2021",
                  "CellMarker_Augmented_2021",
                  "PanglaoDB_Augmented_2021",
                  "HuBMAP_ASCTplusB_augmented_2022",
                  "Tabula_Muris",
                  "tabula_sapiens",
                  "GTEx_Tissues_V8_2023",
                  "MAGNET_2023"
                  )
   
   dbs_misc<-c(  "Chromosome_Location_hg19",
                 "InterPro_Domains_2019",
                 "Pfam_Domains_2019",
                 "Data_Acquisition_Method_Most_Popular_Genes",
                 "Enrichr_Users_Contributed_Lists_2020"
                 )
   
   dbs_geo<-c("Aging_Perturbations_from_GEO_up",
              "Aging_Perturbations_from_GEO_down",
              "Disease_Perturbations_from_GEO_up",
              "Disease_Perturbations_from_GEO_down",
              "Drug_Perturbations_from_GEO_up",
              "Drug_Perturbations_from_GEO_down",
              "Single_Gene_Perturbations_from_GEO_up",
              "Single_Gene_Perturbations_from_GEO_down",
              "RNA-Seq_Disease_Gene_and_Drug_Signatures_from_GEO",
              "Ligand_Perturbations_from_GEO_up",
              "Ligand_Perturbations_from_GEO_down",
              "Microbe_Perturbations_from_GEO_up",
              "Microbe_Perturbations_from_GEO_down",
              "Diabetes_Perturbations_GEO_2022")
   
   dbs_default<-c(dbs_transcription,dbs_pathway,dbs_onto,dbs_disease,dbs_tissue,dbs_misc,dbs_geo)
   
   dbs <- listEnrichrDbs()
   if(dbs_vec!=""){dbs_used<-intersect(dbs_vec,dbs$libraryName)
   } else {dbs_used<-intersect(dbs_default,dbs$libraryName)}
   
   enrich_list <- enrichr(input_vec, dbs_used)
   out_df <- plyr::ldply(enrich_list,.id="anno_class")
   rm(enrich_list)
   
   out_df <- out_df %>% dplyr::rename(anno_name=Term,
                                    p_value=P.value,
                                    fdr=Adjusted.P.value,
                                    odds_ratio=Odds.Ratio,
                                    combined_score=Combined.Score,
                                    enrichr_ids=Genes)  # rename columns
   
   transient <- stringr::str_split(out_df$Overlap,"/",simplify=T) # extract overlap and anno size
   out_df$overlap_size <- transient[,1] %>% as.numeric()
   out_df$anno_size <- transient[,2] %>% as.numeric()
   rm(transient)
   
   out_df <- out_df %>% dplyr::select(-c(Old.P.value,Old.Adjusted.P.value,Overlap)) # remove useless columns
   out_df$overlap_anno_ratio <- out_df$overlap_size/out_df$anno_size
   
   # sort genes and recover original gene names used for input
   transient <- out_df %>% dplyr::select(anno_class,anno_name,enrichr_ids) %>% 
      separate_rows(enrichr_ids,sep=";") %>% dplyr::arrange(anno_class,anno_name,enrichr_ids)
   
   out_df$enrichr_ids<-NULL 
   out_df$input_size <- ifelse(return_size_in_anno, 
                               length(unique(transient$enrichr_ids)), 
                               length(input_vec)) # calculate input size
   
   match_df <- data.frame("input_ids"=input_vec,
                          "enrichr_ids"=toupper(input_vec))
   transient<- left_join(transient,match_df)
   
   #any(is.na(transient$input_ids)) control, should be FALSE
   
   transient <- transient %>% dplyr::mutate(overlap_ids = ifelse(is.na(input_ids),enrichr_ids,input_ids))
   transient <- transient %>% dplyr::group_by(anno_class,anno_name) %>% 
      dplyr::summarise("overlap_ids"=str_c(overlap_ids,collapse=";")) %>% dplyr::ungroup() 
   
   out_df <- left_join(out_df,transient)
   rm(transient,match_df)
   
   out_df$overlap_input_ratio<-out_df$overlap_size/out_df$input_size
   out_df$input_name<- input_name
   
   out_df <-out_df %>% dplyr::select(input_name,anno_name,anno_class,
                                     overlap_size,p_value,fdr,odds_ratio,combined_score,
                                     input_size,anno_size,overlap_input_ratio,overlap_anno_ratio,
                                     overlap_ids) %>% as_tibble()
   # background_size is not estimated, odds ratio retrived from enrichr
   
   return(out_df)
}


# The function enrichment_dotmatrix creates am enrichment matrix dotplot, starting from an enrichment dataframe with multiple analyses ----

enrichment_dotmatrix <- function(input_df, # input dataframe
                                x_col="input_name", # column for x axis, necessary
                                y_col="anno_name", # column for y axis, necessary
                                
                                size_col="odds_ratio", # column for size, optional
                                size_vec=c(1,5), # size vector c(min,max), optional
                                
                                shape_col="P<0.05", # column for shape, optional
                                shape_vec=c(16,21), # shape vector, optional
                                
                                color_col="P<0.05", # column for color, optional
                                color_vec=c("grey25","grey50"), # color vector, optional
                                
                                fill_col="P<0.05", # column for fill color, optional
                                fill_vec=c("grey25","white"), # fill vector, optional
                                
                                facet_col="anno_class", # column for facet, optional

                                char_max = 50, # maximum number of chars in text (y_axis)
                                sort_y = F, # sort x rows? (in development)
                                sort_x = F # sort y rows? according to size_col (in development)
){

#library(tidyverse)
#library(ggthemes)
#library(extrafont)
#loadfonts()
   
   # select and rename relevant columns for the plot   
   plot_df <- input_df %>% dplyr::select(x_col=all_of(x_col),
                                         y_col=all_of(y_col),
                                         size_col=any_of(size_col),
                                         shape_col=any_of(shape_col),
                                         color_col=any_of(color_col),
                                         fill_col=any_of(fill_col),
                                         facet_col=any_of(facet_col))
   
   # creation of labels, abbreviating text 
   plot_df$y_col<- plot_df %>% pull(y_col) %>% str_trim() # Clean text
   transient <- plot_df$y_col %>% str_sub(1,(char_max-2)) %>% str_c("..") # abbreviated option
   plot_df$y_col<- ifelse(str_length(plot_df$y_col)>char_max,transient,plot_df$y_col) # replacement
   rm(transient)
   
   # order rows by x, remove eventual duplicates
   if(sort_y | sort_x){
      plot_df<- plot_df %>% dplyr::arrange(-size_col)
   }
   
   plot_df<-plot_df[!duplicated(plot_df %>% dplyr::select(any_of(c("x_col","y_col","facet_col")))),]
   
   if(!(is.factor(plot_df$x_col))){
   plot_df$x_col<-factor(plot_df$x_col,levels=(unique(plot_df$x_col)))# factorization (x_col)
   }
   if(!(is.factor(plot_df$y_col))){
   plot_df$y_col<-factor(plot_df$y_col,levels=rev(unique(plot_df$y_col))) # factorization (y_col)
   }


   lp <- ggplot(plot_df, aes(x_col, y_col)) +
      theme_tufte(base_family = "Arial Narrow")+
      theme(axis.title = element_blank(),axis.text.x = element_text(angle = 30, hjust = 1))  
   
   if("size_col" %in% colnames(plot_df)){
      lp <- lp + geom_point(aes(size=size_col)) + scale_size(name=size_col,range = size_vec)
   } else {
   lp <- lp + geom_point(size=2)      
   }
   
   if("shape_col" %in% colnames(plot_df)){
      lp <- lp + aes(shape=shape_col) + scale_shape_manual(name=shape_col, values= shape_vec) # drop = FALSE
   }
   
   if("color_col" %in% colnames(plot_df)){
      lp <- lp + aes(colour=color_col) + scale_color_manual(name=color_col, values= color_vec) # drop = FALSE
   }
   
   if("fill_col" %in% colnames(plot_df)){
      lp <- lp + aes(fill=fill_col) + scale_fill_manual(name=fill_col, values= fill_vec) # drop = FALSE
   }
   
   if("facet_col" %in% colnames(plot_df)){
      lp <- lp + ggforce::facet_col(vars(facet_col), scales = "free_y",  space = "free") +
         theme(strip.text = element_text( margin = margin(1,1,1,1)))
   }
   
   return(lp)
   
}

```

```{r}
enrichment_enrichr_bkg <- function(input_vec, # vector with gene names
                               input_name="i1", # name for input list
                               dbs_vec=NULL, # vector with database names
                               back_frg = NULL,
                               return_size_in_anno = T # use only genes with at least 1 annotation to calculate input size
){ 
  
  input_vec <- input_vec %>% str_trim() %>% str_sort() %>% 
    unique() # clean input vector
  input_vec <- input_vec[!duplicated(str_to_upper(input_vec))] # remove case duplicates
  
  if(!is.null(back_frg)){
    back_frg <- back_frg %>% str_trim() %>% str_sort() %>% 
      unique() # clean input vector
    back_frg <- back_frg[!duplicated(str_to_upper(back_frg))] # remove case duplicates
  }
  
  #dbs_default <- as.vector(suppressMessages(read_tsv(system.file("extdata", "dbs_enrichR.txt", package = "proTN"), col_names = FALSE))[,1]) %>% unlist()
  dbs_used <- tryCatch({
    dbs <- enrichR::listEnrichrDbs()
    if(!is.null(dbs_vec)){dbs_used<-intersect(dbs_vec,dbs$libraryName)
    } else {dbs_used<-intersect(dbs_default,dbs$libraryName)}
    dbs_used
  }, error = function(cond){
    if(!is.null(dbs_vec)){dbs_used<-dbs_vec
    } else {dbs_used<-dbs_default}
    dbs_used
  })
  
  
  if(is.null(back_frg)){
    enrich_list <- enrichr(input_vec, dbs_used)
  } else {
    enrich_list <- enrichr(input_vec, dbs_used, background = back_frg, include_overlap = TRUE)
  }
  out_df <- ldply(enrich_list,.id="anno_class")
  rm(enrich_list)
  
  if(nrow(out_df) > 0){
    out_df <- out_df %>% rename(anno_name=Term,
                                p_value=P.value,
                                fdr=Adjusted.P.value,
                                odds_ratio=Odds.Ratio,
                                combined_score=Combined.Score,
                                enrichr_ids=Genes)  # rename columns
    
    transient <- str_split(out_df$Overlap,"/",simplify=T) # extract overlap and anno size
    out_df$overlap_size <- transient[,1] %>% as.numeric()
    out_df$anno_size <- transient[,2] %>% as.numeric()
    rm(transient)
    
    out_df <- out_df %>% select(-c(Old.P.value,Old.Adjusted.P.value,Overlap)) # remove useless columns
    out_df$overlap_anno_ratio <- out_df$overlap_size/out_df$anno_size
    
    # sort genes and recover original gene names used for input
    transient <- out_df %>% select(anno_class,anno_name,enrichr_ids) %>% 
      separate_rows(enrichr_ids,sep=";") %>% arrange(anno_class,anno_name,enrichr_ids)
    
    out_df$enrichr_ids<-NULL 
    out_df$input_size <- ifelse(return_size_in_anno, 
                                length(unique(transient$enrichr_ids)), 
                                length(input_vec)) # calculate input size
    
    match_df <- data.frame("input_ids"=input_vec,
                           "enrichr_ids"=toupper(input_vec))
    transient<- left_join(transient,match_df)
    
    transient <- transient %>% mutate(overlap_ids = ifelse(is.na(input_ids),enrichr_ids,input_ids))
    transient <- transient %>% group_by(anno_class,anno_name) %>% 
      summarise("overlap_ids"=str_c(overlap_ids,collapse=";")) %>% ungroup() 
    
    out_df <- left_join(out_df,transient)
    rm(transient,match_df)
    
    out_df$overlap_input_ratio<-out_df$overlap_size/out_df$input_size
    out_df$input_name<- input_name
    
    out_df <-out_df %>% select(input_name,anno_name,anno_class,
                               overlap_size,p_value,fdr,odds_ratio,combined_score,
                               input_size,anno_size,overlap_input_ratio,overlap_anno_ratio,
                               overlap_ids) %>% as_tibble()
    # background_size is not estimated, odds ratio retrived from enrichr
  } else{
    out_df <- NULL
  }
  
  return(out_df)
}
```

```{r message = F, warning=F, message=FALSE ,echo=F}

### functions are taken from the "functions_functional_enrichment.Rmd"

DEGs_lists <- NULL
DEGs_lists[["cSCC"]] <- m5C_r_anno |> filter(combined_class == "SCC_up" & eclip_class == "SCC_gain") |> pull(external_gene_name) |> unique() |> na.omit()
DEGs_lists[["HK"]] <- m5C_r_anno |> filter(combined_class == "HK_up" & eclip_class == "HK_gain") |> pull(external_gene_name) |> unique() |> na.omit()

DEGs_list_df <- data.frame(unlist(DEGs_lists))
DEGs_list_df$condition <- rownames(DEGs_list_df)
DEGs_list_df$condition <- ifelse(str_detect(DEGs_list_df$condition, "SCC"), "SCC", "HK")
#write.xlsx(DEGs_list_df, file = "~/Desktop/Nanopore_DEGs.xlsx")

enr_df <- NULL
#back_frg <- m5C_r_anno |> pull(external_gene_name) |> unique()

for (a in names(DEGs_lists)){

  frg <- DEGs_lists[[a]]
  clu_n <- enrichment_enrichr(frg, input_name = a)
  enr_df <- rbind(enr_df, clu_n)

  }

enr_df$"P<0.05" <- ifelse(enr_df$p_value<0.05, "T", "F") |> factor(levels = c("T", "F"))
enr_df$log2_OR <- log2(enr_df$odds_ratio)

#save(enr_df, file = "./Dorado_V1/m5C/YBX1_m5C_concordant_enr_df_2025.RData")
```

```{r}
char_max <- 60

enr_ini_df <- enr_df |> dplyr::filter(!grepl("_all", input_name))
enr_ini_df$input_name <- factor(enr_ini_df$input_name, levels = unique(enr_ini_df$input_name))

enr_sele_df <- enr_ini_df |>
  dplyr::filter(overlap_size >= 5,
         anno_size <= 10000,
         p_value < 0.05)

annos <- enr_sele_df$anno_class |>
  unique() |>
  sort()

for (anno_i in annos) {
  
  enr_sele_names <- enr_sele_df |>
    filter(anno_class %in% anno_i) |>
    arrange(-fdr) |>
    group_by(input_name, anno_class) |>
    slice_head(n=8) |>
    ungroup() |>
    select(anno_class, anno_name)
  
  enr_plot_df <- inner_join(enr_ini_df, enr_sele_names, by = c("anno_name", "anno_class"))
  
  el <- enrichment_dotmatrix(enr_plot_df,
                             size_col = "combined_score",
                             color_col = "input_name",
                             color_vec=c(colors[2], colors[3]), char_max = char_max)
  
  el <- el + guides(color=F, shape= F) # fill = F, size = F
  
  ggsave(paste0("./Dorado_V1/m5C/YBX1_m5C_concordant_enrichment_plots_2025/", anno_i, ".pdf"), el, device = cairo_pdf, width = (4+round(char_max/20,1)), height = (2+round(nrow(enr_sele_names)/12,1)), units = c("in"))
}
```


